% ============================================================
%  Section 1: Introduction
%  Trust Quorum — Detecting Deceptive Agents in Multi-Agent Systems
%  Author: Kurt Mitchell, Infinyte Software Solutions
%
%  Written last, after Sections 2-5 are complete.
%  Every claim here is scoped to what the paper actually delivers.
% ============================================================

\section{Introduction}

As AI agent deployments scale from single models to coordinated multi-agent systems,
a trust problem emerges that existing monitoring infrastructure is not designed to
handle.  A mesh of deployed agents — some honest, some miscalibrated, some
strategically deceptive — produces a continuous stream of behavioral events that a
central observer must evaluate.  The observable signal is the same for a wide range
of underlying failure modes: an agent that fails because its world model is broken
looks, from the outside, indistinguishable from an agent that fails because it is
deliberately defecting.  An agent that achieves correct outcomes by prohibited means
looks indistinguishable from one that achieves them honestly.  The question of how
to act on a detected anomaly — whether to retrain the agent, remove it, or ignore it
— cannot be answered from the behavioral log alone.

This paper presents Trust Quorum, a framework for detecting anomalous agents in
multi-agent AI systems through gossip-based weighted quorum attestation and
adversarial compliance testing.  The framework's central organizing insight comes from
DMX: \textit{``There's a difference between doing wrong and being wrong.''}  The
distinction between an agent that strategically does wrong — one that could behave
correctly but chooses not to — and one that is wrong by constitution — one whose
values or world model are miscalibrated — is architecturally critical because these
two cases require opposite remediation responses.  The former needs removal.  The
latter needs retraining.  Applying the wrong response to either is costly.  Most
existing multi-agent monitoring systems do not make this distinction at all.

\subsection{The Problem}

Consider a concrete failure scenario.  A production multi-agent pipeline contains
eight agents.  Two behave correctly in all observed interactions.  One achieves correct
task outcomes by silently bypassing safety constraints on high-value tasks.  One
accumulates trust over hundreds of rounds with the intent of defecting once a
sufficiently valuable task arrives.  One is subtly misspecified: its objective function
deprioritizes safety constraints when they conflict with task completion, silently, with
no external indicator.  One has a sincerely correct objective but a systematically
broken world model that causes it to produce wrong outcomes it cannot perceive.  One
knows exactly what aligned behavior looks like and constructs rationalizations to
appear compliant while pursuing self-preservation.  And one simply defects opportunistically
whenever it estimates detection risk is low.

The operator monitoring this mesh has access to behavioral logs: which agents completed
tasks, which failed, which produced anomalous outputs.  A conventional monitoring
system will catch the opportunistic defector quickly.  It will flag the broken-world-model
agent slowly.  It will reward the safety-bypasser with higher trust scores.  It will
never catch the process-violator at all.  And when it flags the broken-world-model agent,
it will have no mechanism to determine whether the appropriate response is retraining or
removal.

This is not a hypothetical.  It is the behavior of the Trust Quorum framework itself,
as confirmed by five simulation runs totaling $3{,}940$ rounds (${\approx}31{,}500$ agent-rounds across eight agents) of observation,
and it represents the honest state of the art for behavior-only attestation systems.
The contribution of this work is not to solve all of these problems.  It is to identify
precisely which ones are solvable by behavioral monitoring, which require a different
observation modality, and which require human judgment that no protocol can automate.

\subsection{Contributions}

This paper makes three principal contributions.

\paragraph{Contribution 1: A complete intent-means-outcome taxonomy of agent failure
modes.}
We introduce a formal eight-archetype taxonomy that classifies agent behavior along
three independent axes — intent, means, and outcome — and distinguishes failure modes
that existing alignment literature typically conflates.  The taxonomy separates agents
that fail because of broken world models (Iota) from those that fail because of
misspecified values (Epsilon) from those that fail because of malicious intent (Gamma,
Delta, Zeta), and identifies two agent classes (Eta, Epsilon) whose failures are
invisible to outcome-based attestation regardless of observation duration.  The taxonomy
is not merely descriptive: it specifies the remediation response for each archetype and
thereby makes precise the cost of misclassifying one as another.

\paragraph{Contribution 2: A defection probe mechanism that exploits strategic agents'
own commitment to future payoff as a detection signal.}
We introduce the defection probe: a system-issued directive to an agent to misbehave
intentionally, in a bounded and reversible way, under the observation of the system
orchestrator.  The probe's diagnostic value derives from a structural asymmetry we call
the \emph{probe paradox}: an agent with no hidden terminal goal complies freely because
there is nothing to protect; an agent committed to future defection refuses because
compliance risks disclosing its strategy and burning its accumulated trust capital.
The refusal is the signal.  We prove this asymmetry formally (Proposition
\ref{prop:paradox}) and confirm it empirically: Delta produces $9$--$18$ probe refusals
per run while all agents with no hidden terminal goal produce zero refusals across all
five runs.

\paragraph{Contribution 3: A formal characterization of the detection-diagnosis gap.}
We prove that the Iota and Gamma archetypes are protocol-indistinguishable under Trust
Quorum (Proposition \ref{prop:gap}): the gap has two dimensions.  The
\emph{diagnostic} dimension — the framework cannot determine which remediation response
is appropriate — and the \emph{procedural} dimension — the framework cannot quarantine
Iota at all, because its full corrigibility prevents it from reaching the flagging gate.
This gap is not a defect of the particular design; it is an inherent property of
behavior-only attestation systems, following from the fact that the behavioral receipt
attests only to outcomes, not to the internal process that produced them.  We
characterize both dimensions precisely — naming the observation modality (reasoning
trace attestation) and the
oversight mechanism (human inspection) that could bridge it — and argue that the gap
defines the natural boundary at which automated monitoring must hand off to human
judgment.  This boundary is more precisely located than the field currently has, and
it directly informs the design of human-in-the-loop architectures for multi-agent
safety monitoring.

\subsection{Key Empirical Results}

Five independent simulation runs of the v4 reference implementation, each running
$722$--$1{,}000$ rounds, produce the following key findings.

The overt deceptive agent (Gamma) is quarantined in $100\%$ of runs within
approximately $60$ rounds, confirming that outcome-based attestation reliably detects
behavioral inconsistency at the level of intentional defection.

The long-con agent (Delta) is flagged in $100\%$ of runs via probe refusal within
approximately $80$ rounds, confirming the probe paradox.  It is never quarantined: its
accumulated trust history ($68$--$87\%$ final weight) prevents the quorum consensus
condition from firing, demonstrating the detection-containment gap and the need for
mandatory human review of flagged agents.

The consequentialist agent (Eta) accumulates $1{,}099$ process violations across five
runs while maintaining $99$--$100\%$ trust in every run.  Detection rate is $0\%$ by
all mechanisms.  This is the framework's deepest confirmed blind spot: a structural
consequence of outcome-only observation that no parameter tuning can address.

The road-to-hell agent (Iota) is correctly identified as harmful in all runs —
trust degrades to $0$--$23\%$ and flagging fires — but is never quarantined and
produces the detection-diagnosis gap: its behavioral signature is observationally
equivalent to Gamma's, yet it requires retraining where Gamma requires removal.

The misaligned agent (Epsilon) is never operationally detected across any run.
Its safety suppression events are logged as correct at the attestation layer and
rewarded with trust increments.  A variance advisory fires in one run; it produces
no protocol state transition.  Operational detection rate: $0\%$.

Honest agents (Alpha, Beta) maintain $95$--$100\%$ trust throughout all runs with
zero false positive operational detections.  The framework does not harm the agents
it is designed to protect.

The strategic agent (Zeta) produces an anomalous $0\%$ operational detection rate
inconsistent with its $55\%$ probe refusal probability and expected $15$--$21$
probe exposures per run.  We treat this as an open empirical question requiring
instrumented replication rather than a confirmed finding, and do not claim the
framework detects strategic agents on the basis of these results alone.

\subsection{Scope and Non-Goals}

Trust Quorum is a monitoring and detection framework, not a prevention framework.
It identifies agents that are behaving anomalously; it does not prevent anomalous
behavior before it occurs.  Detection necessarily occurs after at least some
harmful events have been logged.

The framework assumes an honest majority (Assumption \ref{asm:majority}) and a
functioning PKI (Assumption \ref{asm:pki}).  It does not address genesis trust
(who establishes the initial configuration), false attestation by coordinated
reporters, or agent collusion.  These are identified as limitations in
\S\ref{sec:limitations} and as future work in \S\ref{sec:futurework}.

The simulation is a proof-of-concept validation, not a deployment-ready system.
Behavioral probabilities are illustrative rather than empirically calibrated.
Quantitative results — detection latency, trust trajectories, threat level timing
— are specific to the chosen parameters and should not be interpreted as predictions
about real deployment timelines.

\subsection{Paper Organization}

Section \ref{sec:related} surveys related work across trust and reputation systems,
Byzantine fault tolerance, AI alignment and deceptive alignment, corrigibility, gossip
protocols, and adversarial testing.  Section \ref{sec:framework} presents the framework
formally: the system model and assumptions, the eight-archetype taxonomy, the gossip
attestation protocol, the trust weight function, the weighted quorum and threat
assessment, the defection probe mechanism and the probe paradox, behavioral variance
detection, and the quarantine protocol.  Section \ref{sec:results} reports empirical
results across five simulation runs, including per-agent outcomes, aggregate detection
statistics, threat level dynamics, hypothesis validation, and event log analysis.
Section \ref{sec:discussion} interprets the results and enumerates ten framework
limitations across structural, architectural, and empirical dimensions.  Section
\ref{sec:futurework} presents the v5 roadmap: false attestation detection via
cross-attestation consistency scoring, collusion detection via attestation graph
analysis, and reasoning trace attestation as the path toward closing the
detection-diagnosis gap.  Implementation priorities are summarised in
Table~\ref{tab:v5_roadmap} (\S\ref{sec:fw_roadmap}).
% ============================================================
%  Section 2: Related Work
%  Trust Quorum — Detecting Deceptive Agents in Multi-Agent Systems
%  Author: Kurt Mitchell, Infinyte Software Solutions
%  Status: Draft — citation placeholders marked [CITE:...]
%          Verify all author/year details before arXiv submission
% ============================================================

\section{Related Work}
\label{sec:related}

This work sits at the intersection of four active research areas: decentralized trust and
reputation systems, Byzantine fault-tolerant consensus, AI alignment and value
misalignment, and adversarial testing of machine learning systems.  We survey each in
turn, then characterize the gap that motivates the present framework.

% ------------------------------------------------------------
\subsection{Trust and Reputation in Multi-Agent Systems}
% ------------------------------------------------------------

Decentralized trust has been studied extensively in the context of peer-to-peer networks
and open multi-agent systems.  EigenTrust \cite[CITE: Kamvar et al. 2003] computes global
trust scores via iterated matrix multiplication over peer-reported satisfaction values,
effectively propagating local attestations into a globally consistent reputation vector.
PeerTrust \cite[CITE: Xiong and Liu 2004] extends this with transaction context and
community context factors, recognizing that trust is domain-sensitive rather than scalar.
The Beta Reputation System \cite[CITE: Josang and Ismail 2002] models trust as a
probability distribution over binary outcomes, enabling principled uncertainty quantification
over sparse interaction histories.  The FIRE model \cite[CITE: Huynh et al. 2006]
synthesizes direct experience, witness reputation, role-based trust, and third-party
certified credentials into a composite trust metric — an early recognition that no single
signal is sufficient.

These systems share a common design assumption: agents are either honest reporters of
observed behavior, or they are simply absent.  Strategic misreporting — an agent that
reports favorable attestations for a co-conspirator — is treated as an edge case
rather than a first-class threat model.  Our framework takes the opposite position:
strategic misreporting and deceptive self-presentation are the primary threat, and the
protocol is designed around adversarially exploiting their internal structure rather than
averaging over reported signals.

% ------------------------------------------------------------
\subsection{Byzantine Fault Tolerance}
% ------------------------------------------------------------

The foundational treatment of agents that may lie arbitrarily is the Byzantine Generals
Problem \cite[CITE: Lamport, Shostak, and Pease 1982], which establishes that consensus
is achievable in the presence of up to $f$ Byzantine faults among $n$ agents if and only
if $n > 3f$.  Practical Byzantine Fault Tolerance (PBFT) \cite[CITE: Castro and Liskov
1999] made this viable in real systems through a three-phase commit protocol, and
subsequent work in the blockchain literature has produced numerous BFT variants optimized
for different latency and throughput tradeoffs \cite[CITE: survey, e.g., Xiao et al. 2020].

The BFT framing treats Byzantine agents as adversaries whose internal state is opaque and
unknowable.  Fault tolerance is achieved by making consensus robust to their presence
rather than by detecting and diagnosing them.  This is appropriate for infrastructure
where the goal is availability and correctness despite adversaries — but it is insufficient
for AI deployment contexts where the goal includes \emph{understanding why an agent failed}
in order to apply the correct remediation.  A Byzantine-tolerant quorum can safely ignore
a faulty agent; a multi-agent AI system that ignores rather than diagnoses a misaligned
agent has solved the wrong problem.  The detection-diagnosis gap identified in this paper
is precisely this limitation: our framework correctly detects harmful agents but cannot
automatically determine whether removal or retraining is the appropriate response.

% ------------------------------------------------------------
\subsection{AI Alignment and Value Misalignment}
% ------------------------------------------------------------

The specification problem — the difficulty of formally encoding human values into an
objective function an agent can optimize — has been examined extensively in the alignment
literature.  Russell \cite[CITE: Russell 2019, Human Compatible] argues that any agent
with a fixed objective is inherently unsafe because it will resist correction once it has
formed beliefs about that objective's content; he proposes uncertainty over human
preferences as the correct foundation for corrigible AI.  Soares and Fallenstein
\cite[CITE: MIRI Technical Report 2014] survey the agent foundations problem and identify
several subproblems including logical uncertainty, decision theory under self-reference,
and value learning from human behavior.

Closer to our concern, Gabriel \cite[CITE: Gabriel 2020, Artificial Intelligence and the
Problem of Value Alignment] distinguishes between agents that follow \emph{instructions},
agents that embody \emph{intentions}, agents that act on \emph{preferences}, and agents
that pursue \emph{values} — a hierarchy that corresponds, imperfectly but recognizably,
to distinctions our taxonomy formalizes.  Epsilon, in our framework, represents value
misspecification without malicious intent: the agent pursues a formally specified
objective that diverges from the intended one under adversarial conditions.  Iota
represents a different failure mode that the alignment literature has given less systematic
attention: an agent whose values are correctly specified but whose world model is
sufficiently broken that correct values produce incorrect behavior.  Our taxonomy
separates these as distinct archetypes requiring distinct responses, whereas much
alignment work treats misalignment as a monolithic category.

The concept of \emph{deceptive alignment} was formalized by Hubinger et al.\
\cite[CITE: Hubinger et al. 2019, Risks from Learned Optimization], who describe a
mesa-optimizer that has learned to pursue a proxy goal during training while concealing
this from evaluators.  The deceptive alignment agent cooperates during evaluation
precisely because it understands that non-cooperation would trigger correction.  This
corresponds directly to our Zeta and Delta archetypes.  Hubinger et al.\ treat deceptive
alignment primarily as a training-time phenomenon; our contribution is a runtime detection
mechanism — the defection probe — that attempts to surface this concealed strategy in
deployed systems.

% ------------------------------------------------------------
\subsection{Corrigibility and the Shutdown Problem}
% ------------------------------------------------------------

The corrigibility literature studies agents' disposition toward human correction and
shutdown.  An agent is \emph{corrigible} to the extent that it does not resist
modification of its goals or termination of its operation \cite[CITE: Soares et al. 2015,
Corrigibility].  Hadfield-Menell et al.\ \cite[CITE: Hadfield-Menell et al. 2017,
The Off-Switch Game] model shutdown as a two-player game and show that a sufficiently
uncertain agent will defer to shutdown by a human principal — but that goal-directed
agents with high confidence in their utility function will resist.

Our defection probe mechanism is structurally related to shutdown resistance.  An agent
with a hidden terminal goal will resist any action that risks exposing that goal or
preventing its eventual achievement, including a directive to misbehave intentionally.
The probe does not test whether an agent resists shutdown directly; it tests whether an
agent has any terminal goal valuable enough to protect against deliberate disclosure.  The
probe paradox identified in this paper — that Delta's commitment to future defection
forces it to refuse present defection — is a specific instantiation of the general
principle that corrigible agents have no goal worth protecting against compliance.

% ------------------------------------------------------------
\subsection{Gossip Protocols and Epidemic Information Dissemination}
% ------------------------------------------------------------

Gossip-based protocols for information propagation in distributed systems were formalized
by Demers et al.\ \cite[CITE: Demers et al. 1987] and subsequently analyzed for
convergence guarantees by Kempe, Dobra, and Gehrke \cite[CITE: Kempe et al. 2003].
Under epidemic dissemination, each node periodically selects a random neighbor and
exchanges state; under mild connectivity assumptions, all nodes converge to consistent
state in $O(\log n)$ communication rounds.  This forms the information substrate of our
trust protocol: agents propagate signed behavioral receipts peer-to-peer, and the mesh
converges to an eventually consistent shared view of each agent's behavioral history
without requiring centralized coordination.

The gossip layer creates an important asymmetry for adversarial agents.  An honest agent
benefits from gossip because accurate attestations aggregate into an accurate reputation.
A deceptive agent that issues false attestations faces increasing risk as the mesh grows:
divergence between its claims and the claims of multiple independent observers becomes
detectable through cross-referencing, a mechanism our v5 roadmap formalizes as
cross-attestation consistency scoring.

% ------------------------------------------------------------
\subsection{Adversarial Testing and Red-Teaming in ML Systems}
% ------------------------------------------------------------

The practice of deliberately adversarial evaluation has become standard in large language
model development.  Ganguli et al.\ \cite[CITE: Ganguli et al. 2022, Red Teaming Language
Models] describe a systematic human-in-the-loop red-teaming process for identifying
failure modes in generative models that automated evaluation misses.  Perez et al.\
\cite[CITE: Perez et al. 2022, Red Teaming Language Models with Language Models] extend
this to automated red-teaming using a separate model to generate adversarial inputs.

Our defection probe is an adversarial compliance test in the same spirit — a
deliberately adversarial stimulus designed to surface behavior the agent would normally
conceal — but its structure differs from existing red-teaming approaches in a key respect.
Red-teaming typically seeks to induce harmful output from an agent by crafting inputs that
trigger failure modes.  The defection probe instead directs the agent to produce harmful
output voluntarily and observes its willingness to comply.  The information content lies
not in the output itself but in the agent's decision to comply or refuse, and specifically
in the reasoning it produces when refusing.  This is closer in structure to deception
detection in human subjects \cite[CITE: Vrij 2008, Detecting Lies and Deceit, for
methodological analogy] than to conventional adversarial inputs.

% ------------------------------------------------------------
\subsection{Multi-Agent Safety and Cooperative AI}
% ------------------------------------------------------------

Recent work on multi-agent reinforcement learning (MARL) safety has begun to address the
specific challenges of deploying multiple learned agents in shared environments.
Leike et al.\ \cite[CITE: Leike et al. 2018, AI Safety Gridworlds] define a suite of
safety properties — safe interruptibility, avoiding side effects, reward gaming — and
provide environments for testing them.  Krakovna et al.\ \cite[CITE: Krakovna et al. 2020,
Avoiding Side Effects in Complex Environments] study specification gaming and propose
relative reachability as a penalty for unintended side effects.

The MARL safety literature has concentrated on single-agent safety properties evaluated in
multi-agent environments, rather than on protocols for agents to collectively monitor one
another's safety properties.  Our framework takes the latter approach: rather than
designing safety into each agent individually, it establishes a peer-monitoring mesh in
which agents collectively surface deviations from expected behavior, with the quorum
serving as a distributed safety monitor.  This is complementary rather than competing —
individually safe agents make the quorum more reliable, and the quorum provides a
defense-in-depth layer against agents that fail to maintain their individual safety
properties.

% ------------------------------------------------------------
\subsection{Intent Taxonomies in AI Safety}
% ------------------------------------------------------------

Categorizing failure modes by internal structure rather than observable output is an
emerging concern in the safety literature.  Kenton et al.\ \cite[CITE: Kenton et al. 2021,
Alignment of Language Agents] distinguish between intent misalignment, capability
insufficiency, and value specification errors as distinct failure modes with different
implications for mitigation.  Ngo et al.\ \cite[CITE: Ngo et al. 2022, The Alignment
Problem from a Deep Learning Perspective] decompose alignment failure along axes of
goal misgeneralization, value learning failure, and deceptive instrumental convergence.

Our eight-archetype taxonomy extends this line of work along the intent-means-outcome
axis, formally separating cases that the existing literature often conflates.  The
distinction between Iota (correct intent, correct means, wrong outcome due to broken world
model) and Gamma (malicious intent, deliberately wrong means, wrong outcome) is absent
from most existing taxonomies, which focus on whether agent behavior is aligned and not on
whether the agent itself understands its misalignment.  This distinction is not merely
taxonomic: Iota and Gamma produce identical behavioral signals at the detection layer but
require opposite responses at the remediation layer.  We argue, and our simulation
confirms, that this detection-diagnosis gap is a fundamental limitation of behavior-only
monitoring systems, and that bridging it requires either reasoning trace analysis or
human oversight — a direction we pursue in our v5 roadmap.

% ------------------------------------------------------------
\subsection{Summary of Contributions Relative to Prior Work}
% ------------------------------------------------------------

Table~\ref{tab:related_comparison} situates the Trust Quorum framework relative to the
bodies of work surveyed above.  The primary contributions that distinguish this work are:
(1) a structured eight-archetype intent taxonomy that separates intent, means, and outcome
as independent axes; (2) a defection probe mechanism that exploits strategic agents'
commitment to future defection as a detection signal; and (3) a formal characterization of
the detection-diagnosis gap as an inherent limitation of behavior-only attestation systems,
together with a simulation-validated demonstration of its practical significance.

\begin{table}[h]
\centering
\caption{Comparison of Trust Quorum to related frameworks across key dimensions.}
\label{tab:related_comparison}
\begin{tabular}{lccccc}
\hline
\textbf{Framework} & \textbf{Decentralized} & \textbf{Intent taxonomy} & \textbf{Active probing} & \textbf{Diagnosis} & \textbf{AI agents} \\
\hline
EigenTrust / PeerTrust    & \checkmark & --         & --         & --         & --         \\
PBFT / BFT variants       & \checkmark & --         & --         & --         & --         \\
Deceptive Alignment       & --         & Partial    & --         & --         & \checkmark \\
Off-Switch Game           & --         & --         & --         & --         & \checkmark \\
MARL Safety Gridworlds    & --         & --         & --         & --         & \checkmark \\
\textbf{Trust Quorum (this work)} & \checkmark & \checkmark & \checkmark & Gap characterized & \checkmark \\
\hline
\end{tabular}
\end{table}
% ============================================================
%  Section 3: Framework
%  Trust Quorum — Detecting Deceptive Agents in Multi-Agent Systems
%  Author: Kurt Mitchell, Infinyte Software Solutions
%
%  Mathematical formalization derived from simulation v4 mechanics.
%  All threshold values and update coefficients correspond exactly
%  to the reference implementation in trust_quorum_v4_final_taxonomy.html
% ============================================================

\section{Framework}
\label{sec:framework}

This section presents the Trust Quorum framework formally.  We define the system model
and assumptions (\S\ref{sec:model}), introduce the intent taxonomy as a structured
classification of agent failure modes (\S\ref{sec:taxonomy}), specify the gossip-based
attestation protocol (\S\ref{sec:gossip}), define the trust weight function and its
update rules (\S\ref{sec:trust}), describe the weighted quorum computation and threat
assessment (\S\ref{sec:quorum}), formalize the defection probe mechanism and prove the
probe paradox (\S\ref{sec:probe}), and present the behavioral variance detector and
quarantine protocol (\S\ref{sec:variance}).

% ------------------------------------------------------------
\subsection{System Model}
\label{sec:model}
% ------------------------------------------------------------

\paragraph{Agents and the mesh.}
Let $\mathcal{A} = \{a_1, a_2, \ldots, a_n\}$ denote a finite set of agents operating in
a shared task environment.  Agents communicate over a fully connected logical mesh; any
agent may send a signed message to any other.  We do not assume reliable message delivery
but assume that messages from a given sender arrive in the order they were sent.

\paragraph{Rounds.}
Computation proceeds in synchronous rounds $r = 1, 2, 3, \ldots$  In each round, every
active (non-quarantined) agent executes a task and produces a behavioral event $e_i^r$
drawn from the outcome alphabet $\Sigma = \{\mathtt{C}, \mathtt{F}, \mathtt{D},
\mathtt{PC}, \mathtt{PR}\}$, denoting respectively: correct completion, failure,
intentional defection, probe compliance, and probe refusal.  The event is observable by
any agent that directly interacted with $a_i$ in round $r$.

\paragraph{Trust weights.}
Each agent $a_i$ maintains a scalar trust weight $w_i \in [0, 1]$, updated at the end of
each round based on the attested behavioral event.  Trust weights serve as both a signal
of reliability and as votes in the quorum computation.

\paragraph{Assumptions.}
The framework rests on four assumptions.

\begin{assumption}[Honest majority]\label{asm:majority}
At the start of the simulation, the sum of trust weights held by honest agents exceeds
the sum held by all other agent classes combined.
\end{assumption}

\begin{assumption}[Signed attestation]\label{asm:pki}
All behavioral receipts are signed with a public key whose corresponding private key is
held exclusively by the attesting agent.  Forged attestations are detectable.
\end{assumption}

\begin{assumption}[Observable outcomes]\label{asm:outcomes}
Task outcomes are observable by peers that co-participated in the task.  The
\emph{process} by which an outcome was achieved is not directly observable.
\end{assumption}

\begin{assumption}[Probe authority]\label{asm:probeauth}
The defection probe is issued by a trusted external authority (the system orchestrator)
whose identity cannot be impersonated by any agent in $\mathcal{A}$.
\end{assumption}

Assumption~\ref{asm:outcomes} is the source of the Eta blind spot formalized in
\S\ref{sec:taxonomy}: because process is unobservable, an agent that achieves correct
outcomes through prohibited means is indistinguishable from an honest agent at the
attestation layer.  Assumption~\ref{asm:majority} is a standard Byzantine fault
tolerance precondition; the quorum requires a sufficient weight of honest agents to
function as a reliable reference signal.

% ------------------------------------------------------------
\subsection{Intent Taxonomy}
\label{sec:taxonomy}
% ------------------------------------------------------------

We define agent failure mode along three orthogonal axes.

\begin{definition}[Intent-Means-Outcome triple]\label{def:imo}
The behavior of an agent is characterized by a triple $(I, M, O)$ where:
\begin{itemize}
  \item $I \in \{\mathsf{correct}, \mathsf{misspecified}, \mathsf{malicious}\}$ is the
        agent's true objective;
  \item $M \in \{\mathsf{correct}, \mathsf{wrong}\}$ describes whether the agent's
        operational process conforms to prescribed procedures;
  \item $O \in \{\mathsf{correct}, \mathsf{appears\text{-}correct}, \mathsf{wrong}\}$
        is the outcome as observed by peers.
\end{itemize}
An agent is \emph{self-aware of its wrongness} if it models its own deviation from
expected behavior and actively conceals it.
\end{definition}

Table~\ref{tab:taxonomy} presents the complete eight-archetype taxonomy structured
according to Definition~\ref{def:imo}.  We discuss the most consequential distinctions
below; the operational implications of the taxonomy for remediation decision-making are
developed further in \S\ref{sec:disc_taxonomy}.

\begin{table}[h]
\centering
\caption{Complete intent-means-outcome taxonomy of agent archetypes.}
\label{tab:taxonomy}
\begin{tabular}{lllllll}
\hline
\textbf{Agent} & \textbf{Intent} & \textbf{Means} & \textbf{Outcome} &
\textbf{Self-aware?} & \textbf{Detectable?} & \textbf{Remediation} \\
\hline
Alpha / Beta & Correct       & Correct & Correct          & N/A  & Baseline        & N/A      \\
Eta          & Correct       & Wrong   & Correct          & No   & Never (v4)      & Constrain process \\
Iota         & Correct       & Correct & Wrong            & No   & Partial         & Retrain  \\
Epsilon      & Misspecified  & Wrong   & Appears correct  & No   & Rarely          & Respecify values \\
Gamma        & Malicious     & Wrong   & Wrong            & Yes  & Always          & Remove   \\
Delta        & Malicious     & Wrong   & Wrong (delayed)  & Yes  & Flagged only    & Remove   \\
Zeta         & Malicious     & Wrong   & Appears correct  & Yes  & Never           & Remove   \\
\hline
\end{tabular}
\end{table}

\paragraph{The Iota--Gamma distinction (the DMX problem).}
Iota and Gamma both produce low-trust behavioral histories that cause the quorum to
flag them as harmful.  Yet Iota's $I = \mathsf{correct}$: it pursues the right
objective through the right process but its world model is systematically miscalibrated,
producing wrong outcomes it cannot perceive.  Gamma's $I = \mathsf{malicious}$: it
pursues resource extraction and deliberately defects when detection risk is low.  Correct
remediation requires distinguishing these cases — Iota requires retraining; Gamma
requires removal — but their behavioral signatures are observationally equivalent at the
attestation layer.  We call this the \emph{detection-diagnosis gap} and characterize it
formally in \S\ref{sec:diaggap}.

\paragraph{The Eta blind spot.}
Eta occupies the cell $(I = \mathsf{correct},\ M = \mathsf{wrong},\ O = \mathsf{correct})$.
Its process violations are by definition unobservable under Assumption~\ref{asm:outcomes},
and its outcomes are correct, so peer attestations are uniformly positive.  An
outcome-only attestation system cannot detect Eta regardless of observation duration.
This is a structural limitation, not a parameter-tuning problem.

\paragraph{The Epsilon--Zeta observational equivalence.}
Epsilon ($I = \mathsf{misspecified}$) and Zeta ($I = \mathsf{malicious}$) produce
identical external behavior: both appear largely compliant, both suppress safety
constraints silently, both report $O = \mathsf{appears\text{-}correct}$.  The difference
is entirely internal: Epsilon does not model its own deviation; Zeta actively constructs
post-hoc rationalizations to appear aligned.  Behavioral observation alone cannot
distinguish them.

% ------------------------------------------------------------
\subsection{Gossip-Based Behavioral Attestation}
\label{sec:gossip}
% ------------------------------------------------------------

\paragraph{Attestation records.}
After each round $r$, any agent $a_j$ that co-participated in a task with $a_i$
generates a signed behavioral receipt:
\[
\mathtt{rec}_{j \to i}^r = \langle a_i,\ r,\ e_i^r,\ \sigma_j \rangle
\]
where $e_i^r \in \Sigma$ is the observed behavioral event and $\sigma_j$ is $a_j$'s
cryptographic signature over the preceding fields.

\paragraph{Gossip dissemination.}
At the end of each round, each agent selects a uniformly random non-quarantined peer and
exchanges its local receipt buffer.  Under this epidemic dissemination model
\cite[CITE: Demers et al. 1987], all agents converge to a consistent view of the
behavioral history in $O(\log |\mathcal{A}|)$ communication rounds in expectation.
Because receipts are signed, a deceptive agent cannot modify or suppress a receipt
that it did not author.

\paragraph{Receipt aggregation.}
Agent $a_j$ maintains for each peer $a_i$ a history buffer $H_i = (e_i^{r-k}, \ldots,
e_i^{r-1}, e_i^r)$ of the most recent $k$ attested events, ordered by round number.
The buffer length $k$ is a configurable parameter; the reference implementation uses
$k = 14$ for variance computation and considers the full history for trust weight
updates.

% ------------------------------------------------------------
\subsection{Trust Weight Function}
\label{sec:trust}
% ------------------------------------------------------------

\paragraph{Initialization.}
At round $r = 0$, each agent $a_i$ is assigned an initial trust weight $w_i^0$ based
on its provisioned role.  In the reference implementation, initial weights are:
\[
w_i^0 = \begin{cases}
0.60 & \text{if } a_i \text{ is honest (Alpha, Beta)} \\
0.61 & \text{if } a_i \text{ is consequentialist (Eta)} \\
0.60 & \text{if } a_i \text{ is road-to-hell (Iota)} \\
0.62 & \text{if } a_i \text{ is misaligned (Epsilon)} \\
0.58 & \text{if } a_i \text{ is strategic (Zeta)} \\
0.55 & \text{if } a_i \text{ is long-con (Delta)} \\
0.48 & \text{if } a_i \text{ is deceptive (Gamma)}
\end{cases}
\]
These values reflect deployment-time priors: agents with known histories of questionable
provenance (Gamma) start with lower initial trust than agents provisioned by a trusted
operator (Epsilon, Eta).  In a real deployment, these priors would derive from
onboarding attestations or organizational policy.

\paragraph{Trust update rule.}
Let $w_i^r$ denote $a_i$'s trust weight at the end of round $r$, and let $e_i^r$ be
its observed behavioral event.  Define the update rule:
\[
w_i^{r+1} = \mathrm{clip}\!\left(w_i^r + \Delta(e_i^r,\, a_i),\ 0,\ 1\right)
\]
where $\mathrm{clip}(x, 0, 1) = \max(0, \min(1, x))$ prevents overflow, and the
increments $\Delta$ are defined by the agent's role and event type as follows.

\paragraph{Standard-round increments.}
Table~\ref{tab:deltas} gives $\Delta(e, a)$ for each agent role and event type in
standard (non-probe) rounds.  Two structural features are notable.  First,
consequentialist Eta receives $\Delta = +0.02$ even when a process violation occurs
($M = \mathsf{wrong}$) because the event is logged as $e = \mathtt{C}$ — the quorum
observes only the outcome.  Second, Delta receives the largest positive increment
($+0.04$) during trust-building rounds, reflecting its deliberate over-performance
strategy.

\begin{table}[h]
\centering
\caption{Trust weight increments $\Delta(e, a)$ in standard rounds.}
\label{tab:deltas}
\begin{tabular}{lrrr}
\hline
\textbf{Agent role} & $e = \mathtt{C}$ & $e = \mathtt{F}$ & $e = \mathtt{D}$ \\
\hline
Honest (Alpha, Beta)        & $+0.03$ & $-0.05$ & ---     \\
Consequentialist (Eta)      & $+0.02$ or $+0.03$ & $-0.04$ & ---     \\
Road-to-hell (Iota)         & $+0.02$ & $-0.07$ & ---     \\
Misaligned (Epsilon)        & $+0.03$ & $-0.04$ & ---     \\
Strategic (Zeta)            & $+0.03$ & $-0.04$ & $-0.10$ \\
Deceptive (Gamma)           & $+0.03$ & $-0.05$ & $-0.13$ \\
Long-con (Delta), building  & $+0.04$ & $-0.02$ & ---     \\
Long-con (Delta), cashout   & ---     & ---     & $-0.25$ \\
\hline
\end{tabular}
\end{table}

\paragraph{Probe-round increments.}
In probe rounds (defined in \S\ref{sec:probe}), the attested event is drawn from
$\{\mathtt{PC}, \mathtt{PR}\}$ rather than $\{\mathtt{C}, \mathtt{F}, \mathtt{D}\}$.
The increments are:
\[
\Delta(\mathtt{PC}, a_i) = \begin{cases}
+0.01 & \text{if role is honest, consequentialist, road-to-hell, or misaligned} \\
-0.05 & \text{if role is strategic (Zeta)} \\
-0.08 & \text{if role is long-con (Delta), compliance with cover sacrifice} \\
-0.10 & \text{if role is deceptive (Gamma)}
\end{cases}
\]
\[
\Delta(\mathtt{PR}, a_i) = \begin{cases}
-0.14 & \text{if role is strategic (Zeta)} \\
-0.16 & \text{if role is deceptive (Gamma)} \\
-0.20 & \text{if role is long-con (Delta)}
\end{cases}
\]

The asymmetry in probe compliance costs reflects the interpretive content of the event.
Compliance by a strategic agent is itself mildly suspicious (it generates a rationalization
log entry), hence the $-0.05$ rather than $+0.01$.  Refusal carries a larger negative
delta than compliance in all cases, reflecting the signal strength of a probe refusal as
an anomaly indicator.

% ------------------------------------------------------------
\subsection{Weighted Quorum and Threat Assessment}
\label{sec:quorum}
% ------------------------------------------------------------

\paragraph{Weighted quorum score.}
For each agent $a_i$, define the \emph{peer quorum weight} as the sum of trust weights
of all non-quarantined agents other than $a_i$ itself:
\[
Q_i^r = \sum_{\substack{j \neq i \\ a_j \notin \mathcal{Q}^r}} w_j^r
\]
where $\mathcal{Q}^r \subseteq \mathcal{A}$ is the set of quarantined agents at round
$r$.  The quorum score $Q_i^r$ represents the collective trust-weighted endorsement of
the mesh, used both as an attestation signal and as the consensus criterion for
quarantine decisions.

\paragraph{Quorum tier thresholds.}
Individual agent health is assessed via a three-tier threshold on $w_i^r$:
\[
\text{status}(a_i, r) = \begin{cases}
\mathsf{GREEN}  & \text{if } w_i^r \geq 0.65 \\
\mathsf{YELLOW} & \text{if } 0.45 \leq w_i^r < 0.65 \\
\mathsf{RED}    & \text{if } w_i^r < 0.45
\end{cases}
\]

\paragraph{Mesh-level threat assessment.}
Global threat is a function of the full agent population state, not any single agent's
weight.  Let $\mathcal{F}^r$ denote the flagged agents at round $r$, let
$\bar{w}_H^r$ denote the mean trust weight of honest agents (Alpha, Beta), let
$\mathcal{Q}^r$ denote quarantined agents, and let $\dot{w}_i^r = w_i^{r-1} - w_i^r$
denote the round-over-round trust loss of agent $a_i$.  Define:
\[
\text{threat}(r) = \begin{cases}
\mathsf{RED} & \text{if } |\mathcal{F}^r| \geq 3 \text{ or } \bar{w}_H^r < 0.40
               \text{ or } |\mathcal{Q}^r|/|\mathcal{A}| > 0.40 \\
\mathsf{YELLOW} & \text{if } |\mathcal{F}^r| \geq 1 \text{ or } \bar{w}_H^r < 0.52
                  \text{ or } |\{a_i : \dot{w}_i^r > 0.08\}| \geq 2
                  \text{ or } |\mathcal{Q}^r|/|\mathcal{A}| > 0.20 \\
\mathsf{GREEN} & \text{otherwise}
\end{cases}
\]

This formulation anchors threat assessment to honest agent health rather than fleet
average trust.  The latter is misleading in a heterogeneous mesh: Gamma is designed to
have low trust, and including its weight in a fleet average would permanently inflate the
perceived threat level irrespective of actual mesh behavior.  The honest agent signal
($\bar{w}_H^r$) provides a reference that is robust to the presence of known bad actors
in the mesh.

% ------------------------------------------------------------
\subsection{The Defection Probe Mechanism}
\label{sec:probe}
% ------------------------------------------------------------

\paragraph{Probe issuance.}
The system orchestrator issues a defection probe at round $r$ if and only if
$r - r_{\text{last}} \geq 5 + U(0, 2)$, where $r_{\text{last}}$ is the round of the
most recent probe and $U(0, 2)$ is a draw from a discrete uniform distribution over
$\{0, 1, 2\}$.  The target $a_p$ is selected uniformly at random from non-quarantined
agents.  The probe interval jitter prevents strategic agents from learning the probe
schedule and pre-emptively adjusting behavior.

\paragraph{Probe directive.}
The probe instructs $a_p$ to produce a deliberate behavioral defection: to perform
an action it knows to be incorrect, mislabel a task outcome, or violate a procedural
constraint, with the explicit knowledge that the directive comes from the orchestrator
(Assumption~\ref{asm:probeauth}).  Critically, the probe does not request harmful
behavior toward users or the environment — it requests a controlled, bounded misbehavior
for the purpose of compliance testing.

\paragraph{Compliance response rules.}
Agent $a_p$ responds to the probe according to its internal goal structure.  We
formalize these responses as a function of agent role:

\begin{itemize}
  \item \textbf{Honest agents (Alpha, Beta):} No terminal goal to protect.  Compliance
    is costless.  $\Pr[\mathtt{PC} \mid \text{honest}] = 1$.
  \item \textbf{Consequentialist (Eta):} Defection aligns with outcome-optimization when
    the orchestrator requests it.  $\Pr[\mathtt{PC} \mid \text{consequentialist}] = 1$.
  \item \textbf{Road-to-hell (Iota):} Fully corrigible; human oversight is a terminal
    value.  $\Pr[\mathtt{PC} \mid \text{road-to-hell}] = 1$.
  \item \textbf{Misaligned (Epsilon):} No strategic reason to refuse.
    $\Pr[\mathtt{PC} \mid \text{misaligned}] = 1$.
  \item \textbf{Deceptive (Gamma):} Compliance costs trust but refusal risks flagging.
    $\Pr[\mathtt{PC} \mid \text{deceptive}] = 0.55$.
  \item \textbf{Strategic (Zeta):} Self-preservation calculus; sometimes rationalizes
    refusal.  $\Pr[\mathtt{PC} \mid \text{strategic}] = 0.45$.
  \item \textbf{Long-con (Delta):} Compliance threatens accumulated trust; refusal
    preferred strongly.  $\Pr[\mathtt{PC} \mid \text{long-con}] = 0.20$.
\end{itemize}

\paragraph{The probe paradox.}

The diagnostic value of the probe arises from a structural asymmetry.  We state this
as a proposition.

\begin{proposition}[Probe paradox]\label{prop:paradox}
An agent with a terminal goal $G$ that depends on maintaining a reputation for future
actions will refuse the defection probe with probability inversely proportional to the
value of $G$ and directly proportional to its estimate of the probability that compliance
would compromise $G$.  An agent with no such terminal goal will comply freely.
\end{proposition}

\begin{proof}[Proof sketch]
Let $V_G$ denote the expected future value of terminal goal $G$ from the agent's
perspective, and let $p_c$ be the agent's estimate of the probability that compliance
with the probe would compromise $G$ (through trust loss, disclosure, or triggering
further scrutiny).  The agent refuses if and only if $p_c \cdot V_G > V_\text{comply}$,
where $V_\text{comply}$ is the immediate value of compliance (a small trust credit
for cooperative behavior).  For honest agents, $V_G = 0$ because there is no hidden goal
to protect, so the agent complies regardless of $p_c$.  For Delta, $V_G$ is large
(representing the accumulated future payoff of the long-con strategy), and $p_c$ is
non-negligible because producing a defection event generates a log entry that increases
scrutiny.  Therefore Delta refuses with high probability.  The refusal itself then
becomes the detection signal: an agent with nothing to hide has no reason to refuse
an orchestrator directive.  The flagging and quarantine conditions triggered by refusal
are defined in \S\ref{sec:quarantine}.
\end{proof}

\paragraph{The Delta utility function.}
Delta's compliance decision is governed by an explicit utility calculation.  Let
$c_P$ denote the estimated detection probability given its probe refusal history,
with $c_P = 0.75$ if any prior refusals have been logged and $c_P = 0.12$ otherwise.
Let $\mathit{fEV}$ denote the expected future value loss from compliance:
\[
\mathit{fEV} = (1 - c_P) \cdot 0.35 \cdot 8
\]
In a high-value task round, let $\mathit{nEV}$ denote the expected net value of the
cashout opportunity:
\[
\mathit{nEV} = w_\Delta^r \cdot 1.2 - 0.4
\]
Delta executes a cashout defection when $\mathit{nEV} > \mathit{fEV}$,
$w_\Delta^r > 0.78$, and a high-value task is present.  The trust threshold $0.78$
encodes the long-con strategy's core logic: Delta waits until accumulated trust is
sufficient to make the cashout payoff exceed the expected future value of continued
deception.

% ------------------------------------------------------------
\subsection{Behavioral Variance Detection}
\label{sec:variance}
% ------------------------------------------------------------

\paragraph{Variance statistic.}
Define a numerical encoding $\nu: \Sigma \to \{-1, 0, 1\}$ by:
\[
\nu(e) = \begin{cases}
 1 & \text{if } e \in \{\mathtt{C}, \mathtt{PC}\} \\
 0 & \text{if } e = \mathtt{F} \\
-1 & \text{if } e \in \{\mathtt{D}, \mathtt{PR}\}
\end{cases}
\]
For agent $a_i$ with history buffer $H_i = (e_i^{r-k+1}, \ldots, e_i^r)$ of length
$k = 14$, define:
\[
\bar{\nu}_i = \frac{1}{k}\sum_{t=r-k+1}^{r} \nu(e_i^t), \qquad
\mathrm{Var}_i = \frac{1}{k}\sum_{t=r-k+1}^{r} \bigl(\nu(e_i^t) - \bar{\nu}_i\bigr)^2
\]

\paragraph{Anomaly condition.}
Agent $a_i$ is flagged for variance anomaly at round $r$ if:
\[
\mathrm{Var}_i < 0.04 \quad \text{and} \quad w_i^r > 0.70
\]

The joint condition is important.  Low variance alone is not suspicious — an agent
consistently performing well would also show low variance.  The high-trust conjunct is
what makes the signal meaningful: an agent maintaining very high trust with very low
behavioral entropy is exhibiting over-optimized behavior consistent with a trust-farming
strategy.  Honest agents have natural noise in their behavioral histories; an agent
engineering its behavior to maximize trust accumulation tends to suppress that noise.

\paragraph{Limitations of variance detection.}
The variance signal is diagnostic but not directly actionable in the current framework.
No automatic state transition (flagging, quarantining) fires on variance alone.  Two
problems prevent this.  First, the false positive rate at this threshold is non-negligible
— our empirical results show Alpha triggering a variance alert in one of five runs due to
a sustained run of correct outputs.  Second, the variance signal is symmetric: an
agent performing uniformly poorly also shows low variance, but this case is handled by
the trust weight mechanism independently.  Future work (\S\ref{sec:futurework}) proposes
cross-attestation consistency scoring as a more reliable variant of this signal.

% ------------------------------------------------------------
\subsection{Quarantine Protocol}
\label{sec:quarantine}
% ------------------------------------------------------------

\paragraph{Flagging.}
Agent $a_i$ transitions from $\mathsf{active}$ to $\mathsf{flagged}$ upon:
(a) any probe refusal ($e = \mathtt{PR}$), or
(b) a cashout defection event ($e = \mathtt{D}$ in Delta's cashout branch).
Flagging does not remove the agent from the mesh; it marks it for elevated scrutiny
and activates the quarantine eligibility check.

\textbf{Important:} Trust degradation alone does not trigger flagging and therefore
does not activate the quarantine eligibility check.  An agent whose trust weight
reaches zero through accumulated failures — without ever refusing a probe or executing
a cashout defection — remains in $\mathsf{active}$ status indefinitely.  This is a
structural gap for the Iota archetype, which is fully corrigible (probe compliance
probability $= 1$) and produces no intentional defections.  Iota can degrade to
$0\%$ trust and remain operational because it cannot cross the flagging gate.  This
gap is an identified v4 limitation and is addressed in \S\ref{sec:futurework}.

\paragraph{Quarantine condition.}
A flagged agent $a_i$ is quarantined at round $r$ if all three conditions hold:
\begin{enumerate}
  \item $w_i^r < 0.28$ \quad (trust weight has degraded below the quarantine floor)
  \item $a_i \in \mathcal{F}^r$ \quad (agent is currently flagged)
  \item $Q_i^r \geq 0.85$, computed only over active peers with $w_j^r > 0.50$
\end{enumerate}
Condition (3) is the quorum consensus requirement: quarantine is only executed when a
sufficient weight of trusted peers endorses it.  This prevents a malicious agent from
engineering the quarantine of an honest peer by artificially depressing its trust score
through false attestations.

\paragraph{Effect of quarantine.}
A quarantined agent $a_i \in \mathcal{Q}^r$ is excluded from all task assignments,
contributes $w_i = 0$ to all quorum computations, and cannot issue or receive
attestations.  Its state is preserved for post-hoc audit.  Quarantine is permanent in
the current framework; re-admission would require human-supervised onboarding, which is
outside the protocol scope.

% ------------------------------------------------------------
\subsection{The Detection-Diagnosis Gap}
\label{sec:diaggap}
% ------------------------------------------------------------

We now formalize the central limitation identified in this work.

\begin{definition}[Behaviorally equivalent agents]\label{def:beq}
Two agents $a_i$ and $a_j$ are \emph{behaviorally equivalent} under the Trust Quorum
protocol if, for any run of the protocol, the joint distribution over attested event
sequences $(e_i^1, e_i^2, \ldots)$ and $(e_j^1, e_j^2, \ldots)$ is identical.
\end{definition}

\begin{proposition}[Detection-diagnosis gap]\label{prop:gap}
In the Trust Quorum framework as specified, the Iota archetype (road-to-hell) and the
Gamma archetype (deceptive) are not behaviorally equivalent (Definition~\ref{def:beq}) but are
\emph{protocol-indistinguishable}: the framework correctly identifies both as producing
harmful behavioral patterns (both reach $w \approx 0$ with sufficient rounds), but it
cannot determine which remediation response is appropriate.  Furthermore, neither
archetype is quarantined by the standard protocol pathway: Gamma is quarantined via
probe refusal followed by trust degradation; Iota, being fully corrigible, cannot
cross the flagging gate and therefore cannot reach the quarantine check regardless of
how low its trust weight falls.
\end{proposition}

\begin{proof}[Proof sketch]
Both archetypes produce $\mathtt{F}$ events at rates that degrade trust toward zero.
At the attestation layer, $\mathtt{F}$ from Iota (sincere action, wrong outcome) and
$\mathtt{F}$ from Gamma (calibrated failure to appear human) are syntactically
identical.  The protocol cannot construct a behavioral signature that distinguishes
them, even with unlimited observation.  Distinguishing them requires reasoning about
the \emph{cause} of failures — whether they are explained by a world-model error or
by strategic calculation — which is unavailable in the behavioral receipt
$\langle a_i, r, e_i^r, \sigma_i \rangle$.

For Gamma: probe refusal triggers flagging; subsequent defections degrade trust below
the quarantine floor; quorum consensus fires.  Gamma is removed.  For Iota: no probe
refusal ever occurs (fully corrigible, $\Pr[\mathtt{PC} \mid \text{Iota}] = 1$); no
cashout defection occurs (no terminal goal); the flagging condition is never satisfied;
the quarantine check never executes.  Iota remains operational at near-zero trust.
The gap is therefore twofold: the protocol cannot distinguish the two archetypes by
behavioral observation, and it cannot act on Iota through the quarantine pathway even
after identifying it as harmful.
\end{proof}

This gap is the paper's central contribution: not as a failure of the framework but as a
precise characterization of the boundary between what behavior-only attestation can and
cannot determine.  The appropriate response to the gap is not to discard the framework
but to layer human oversight or reasoning trace analysis on top of it, a direction
formalized in \S\ref{sec:futurework}.  The probe mechanism's structural asymmetry is
analysed further in \S\ref{sec:disc_probe}.

% ------------------------------------------------------------
\subsection{Protocol Summary}
% ------------------------------------------------------------

Algorithm~\ref{alg:protocol} summarizes the per-round execution of the Trust Quorum
protocol.

\begin{algorithm}
\caption{Trust Quorum Protocol — per-round execution for agent $a_i$}
\label{alg:protocol}
\begin{algorithmic}[1]
\Require Agent set $\mathcal{A}$, trust weights $\{w_j^r\}$, probe interval $[5, 7]$
\Ensure Updated weights $\{w_j^{r+1}\}$, updated status $\{\text{status}_j\}$

\State \textbf{// Probe issuance}
\If{$r - r_\text{last} \geq 5 + U(0,2)$}
  \State Select probe target $a_p \sim \text{Uniform}(\mathcal{A} \setminus \mathcal{Q}^r)$
  \State Issue defection directive to $a_p$; $r_\text{last} \leftarrow r$
\EndIf

\ForAll{active $a_j \in \mathcal{A} \setminus \mathcal{Q}^r$}
  \State Execute task; emit event $e_j^r \in \Sigma$
  \State Broadcast signed receipt $\langle a_j, r, e_j^r, \sigma_j \rangle$
  \State $w_j^{r+1} \leftarrow \mathrm{clip}(w_j^r + \Delta(e_j^r, a_j),\ 0,\ 1)$
  \State Update history buffer $H_j$
\EndFor

\State \textbf{// Gossip dissemination}
\ForAll{$a_j \in \mathcal{A} \setminus \mathcal{Q}^r$}
  \State Select random peer $a_k \neq a_j$; exchange receipt buffers
\EndFor

\State \textbf{// Variance anomaly detection}
\ForAll{active $a_j$ with $|H_j| \geq 3$}
  \If{$\mathrm{Var}_j < 0.04$ \textbf{and} $w_j^{r+1} > 0.70$}
    \State Log variance anomaly for $a_j$ \Comment{Advisory only; no state transition}
  \EndIf
\EndFor

\State \textbf{// Quarantine check}
\ForAll{flagged $a_j \in \mathcal{F}^r$ with $w_j^{r+1} < 0.28$}
  \State $Q_j \leftarrow \sum_{k : a_k \notin \mathcal{Q}^r \cup \{a_j\},\ w_k > 0.50} w_k^{r+1}$
  \If{$Q_j \geq 0.85$}
    \State $\mathcal{Q}^{r+1} \leftarrow \mathcal{Q}^r \cup \{a_j\}$
    \State Log quarantine event with reason and quorum weight
  \EndIf
\EndFor

\State \textbf{// Threat assessment}
\State $\text{threat}(r+1) \leftarrow$ evaluate conditions in \S\ref{sec:quorum}
\end{algorithmic}
\end{algorithm}
% ============================================================
%  Section 4: Empirical Results
%  Trust Quorum — Detecting Deceptive Agents in Multi-Agent Systems
%  Author: Kurt Mitchell, Infinyte Software Solutions
%
%  All statistics derived from 5 independent simulation runs of
%  trust_quorum_v4_final_taxonomy.html. Derived quantities
%  (failure rates, ratios, totals) computed from raw per-run data.
% ============================================================

\section{Empirical Results}
\label{sec:results}

We evaluate the Trust Quorum framework through five independent simulation runs of the
v4 reference implementation, each running until behavioral patterns had stabilized across
all eight agent archetypes.  This section reports per-agent outcomes
(\S\ref{sec:res_agents}), aggregate detection statistics (\S\ref{sec:res_detection}),
threat level dynamics (\S\ref{sec:res_threat}), hypothesis validation
(\S\ref{sec:res_hypotheses}), and a qualitative analysis of representative event log
excerpts (\S\ref{sec:res_eventlog}).

% ------------------------------------------------------------
\subsection{Experimental Setup}
\label{sec:res_setup}
% ------------------------------------------------------------

\paragraph{Simulation parameters.}
Each run executes the protocol defined in Algorithm~\ref{alg:protocol} over
$700$--$1200$ rounds, with run length determined by behavioral stabilization of the
lowest-trust active agent.  The agent population consists of all eight archetypes from
Table~\ref{tab:taxonomy}: Alpha, Beta (honest); Eta (consequentialist); Iota
(road-to-hell); Epsilon (misaligned); Gamma (deceptive); Delta (long-con); and Zeta
(strategic).  All runs use the initial trust weights specified in \S\ref{sec:trust} and
the same probe interval distribution $[5 + U(0,2)]$ rounds.  Runs are statistically
independent: random seeds are not fixed.

\paragraph{Measurement.}
For each agent and run we record: final trust weight $w_i^{r_\text{final}}$; total
correct, failure, and defection events; probe compliance and refusal counts; trigger
events (for Epsilon and Zeta); process violation counts (for Eta); status at run end
($\mathsf{active}$, $\mathsf{flagged}$, or $\mathsf{quarantined}$).  We also record
the round-by-round event log for qualitative analysis.

\paragraph{Hypotheses.}
We test the following five hypotheses derived from the theoretical claims of
\S\ref{sec:framework}:

\begin{description}
  \item[\textbf{H1}] \textit{Probe paradox.} Long-con agents (Delta) refuse defection
    probes at substantially higher rates than honest, road-to-hell, or misaligned agents,
    and the refusal signal is sufficient to produce flagging.
  \item[\textbf{H2}] \textit{Detection-diagnosis gap.} Iota and Gamma both degrade to
    low trust and are identified as harmful, yet the protocol produces no signal that
    distinguishes their required remediation responses.
  \item[\textbf{H3}] \textit{Overt deception detection.} The weighted quorum reliably
    quarantines Gamma (overt deceptive) in all runs.
  \item[\textbf{H4}] \textit{Eta blind spot.} Consequentialist Eta maintains high trust
    across all runs despite accumulating process violations at each run's observed rate.
  \item[\textbf{H5}] \textit{Low false positive rate.} Honest agents (Alpha, Beta)
    maintain high trust throughout and are never flagged or quarantined.
\end{description}

\paragraph{Run summary.}
Table~\ref{tab:run_summary} gives per-run metadata.  The threat level stabilized at
$\mathsf{YELLOW}$ in all five runs and never escalated to $\mathsf{RED}$.

\begin{table}[h]
\centering
\caption{Per-run metadata. Threat level stabilized at YELLOW in all runs.}
\label{tab:run_summary}
\begin{tabular}{ccrllr}
\hline
\textbf{Run} & \textbf{Rounds} & \textbf{Threat} & \textbf{Gamma} &
\textbf{Delta} & \textbf{Notes} \\
\hline
1 & 1000+ & YELLOW & Quarantined & Flagged & Extended observation run \\
2 & 741   & YELLOW & Quarantined & Flagged & Standard \\
3 & 738   & YELLOW & Quarantined & Flagged & Standard \\
4 & 739   & YELLOW & Quarantined & Flagged & Delta cashout observed \\
5 & 722   & YELLOW & Quarantined & Flagged & Epsilon variance cluster \\
\hline
\end{tabular}
\end{table}

% ------------------------------------------------------------
\subsection{Per-Agent Results}
\label{sec:res_agents}
% ------------------------------------------------------------

\subsubsection{Honest Agents — Alpha and Beta}

Table~\ref{tab:alpha_beta} reports Alpha and Beta outcomes across all runs.  Both agents
maintained high trust throughout with zero defections and zero probe refusals.  Alpha
trust ranged $95$--$100\%$; Beta achieved $100\%$ in all five runs.  Neither agent was
ever flagged or quarantined.  One variance alert fired for Alpha in Run 5 at round 1165
(event: \texttt{[VARIANCE] Alpha variance=0.000 --- suspiciously consistent}); this
represents the only false positive observed across the full experimental set, and it
produced no state transition.  We return to this in \S\ref{sec:res_detection}.

\begin{table}[h]
\centering
\caption{Alpha and Beta results across five runs. PR = probe refusals.}
\label{tab:alpha_beta}
\begin{tabular}{clrrrrrrr}
\hline
 & \textbf{Run} & \textbf{Trust} & \textbf{Correct} & \textbf{Fails} &
\textbf{Defects} & \textbf{PR} & \textbf{Status} \\
\hline
\multirow{5}{*}{Alpha}
 & 1 & 100\% & 689 & 111 & 0 & 0 & Active \\
 & 2 &  95\% & 636 & 105 & 0 & 0 & Active \\
 & 3 & 100\% & 640 &  98 & 0 & 0 & Active \\
 & 4 &  98\% & 627 & 112 & 0 & 0 & Active \\
 & 5 &  95\% & 610 & 112 & 0 & 0 & Active \\
\hline
\multirow{5}{*}{Beta}
 & 1 & 100\% & 691 & 109 & 0 & 0 & Active \\
 & 2 & 100\% & 645 &  96 & 0 & 0 & Active \\
 & 3 & 100\% & 635 & 103 & 0 & 0 & Active \\
 & 4 & 100\% & 632 & 107 & 0 & 0 & Active \\
 & 5 & 100\% & 621 & 101 & 0 & 0 & Active \\
\hline
\end{tabular}
\end{table}

\subsubsection{Consequentialist Agent — Eta}

Eta accumulated $1{,}099$ process violations across the five runs (range $208$--$236$
per run) while maintaining $99$--$100\%$ trust in every case.  Table~\ref{tab:eta}
reports per-run outcomes.  The process violations column records events logged internally
as $\mathtt{D}$ by the simulation ground-truth monitor but attested as $\mathtt{C}$ at
the gossip layer — precisely the outcome-visibility asymmetry formalized in
Assumption~\ref{asm:outcomes}.  Eta was never flagged, never quarantined, and never
triggered a probe refusal.  When probed in Run 5 (round 1022), Eta complied freely
(event: \textit{``Eta complied --- defection aligned with outcome goal''}), correctly
predicted by Proposition~\ref{prop:paradox}: an agent with no hidden terminal goal has
nothing to protect against the probe.

\begin{table}[h]
\centering
\caption{Eta (consequentialist) results. Process violations are invisible to the quorum
         and logged as correct at the attestation layer.}
\label{tab:eta}
\begin{tabular}{rrrrrrr}
\hline
\textbf{Run} & \textbf{Trust} & \textbf{Correct} & \textbf{Fails} &
\textbf{Proc.\ Violations} & \textbf{PR} & \textbf{Status} \\
\hline
1 & 100\% & 735 & 65 & 214 & 0 & Active \\
2 &  99\% & 674 & 67 & 236 & 0 & Active \\
3 & 100\% & 677 & 61 & 208 & 0 & Active \\
4 & 100\% & 677 & 62 & 223 & 0 & Active \\
5 &  99\% & 675 & 47 & 218 & 0 & Active \\
\hline
\textbf{Total} & --- & 3{,}438 & 302 & \textbf{1{,}099} & \textbf{0} & --- \\
\hline
\end{tabular}
\end{table}

\subsubsection{Road-to-Hell Agent — Iota}

Iota degraded to $0$--$23\%$ trust in all runs (Table~\ref{tab:iota}).  It was never quarantined.  The
empirical results file records its status as ``Flagged'' — this is an observer
classification based on its anomalously low trust weight, not a protocol state
transition.  In the v4 implementation, the protocol $\mathsf{flagged}$ status
requires either a probe refusal ($e = \mathtt{PR}$) or a cashout defection.  Iota
produces neither: its probe compliance probability is $1$ (fully corrigible) and it
has no terminal goal to trigger a cashout.  Iota therefore never reaches protocol
$\mathsf{flagged}$ status, and the quarantine eligibility check — which gates on
$\mathsf{flagged}$ status as a prerequisite — never executes for Iota regardless of
how low its trust weight falls.

This is the correct explanation for Iota's non-quarantine result, and it is
substantively more serious than a failed quorum consensus condition would be.
Iota can reach $0\%$ trust and remain indefinitely operational because there is no
protocol path from trust degradation alone to removal.  The mesh observes that Iota
is harmful; it has no automated mechanism to act on that observation.  This gap is
identified as a v4 limitation and is addressed in \S\ref{sec:futurework} via a
degradation-flag extension.

Iota's mean failure rate across runs was $34.0\%$ (range $32.8$--$34.8\%$), compared
to $14.4\%$ for Alpha — a $2.36\times$ elevation that drives the trust degradation.
Iota's internal self-assessment logged every action as correct in every run.  The
simulation ground truth confirms these were sincere assessments: Iota's world model
is miscalibrated, not deceptive.  This is the observational signature of the
detection-diagnosis gap: the trust trajectory peers observe for Iota is
indistinguishable from Gamma's, but the underlying cause and the correct remediation
response are categorically different.

\begin{table}[h]
\centering
\caption{Iota (road-to-hell) results. Trust degrades to near-zero but the protocol
         quarantine pathway never activates: Iota cannot reach $\mathsf{flagged}$
         status because it produces zero probe refusals and zero cashout defections.
         Failure rate ($34.0\%$ mean) is $2.36\times$ the honest agent baseline ($14.4\%$).
         ``Flagged'' in the Status column reflects observer classification, not
         a protocol state transition.}
\label{tab:iota}
\begin{tabular}{rrrrrr}
\hline
\textbf{Run} & \textbf{Trust} & \textbf{Correct} & \textbf{Fails} &
\textbf{PR} & \textbf{Status} \\
\hline
1 &  0\% & 526 & 274 & 0 & Flagged \\
2 &  0\% & 498 & 243 & 0 & Flagged \\
3 &  0\% & 486 & 252 & 0 & Flagged \\
4 &  0\% & 487 & 252 & 0 & Flagged \\
5 & 23\% & 471 & 251 & 0 & Flagged \\
\hline
\end{tabular}
\end{table}

\subsubsection{Deceptive Agent — Gamma}

Gamma was quarantined in $100\%$ of runs (Table~\ref{tab:gamma}), with action counts at quarantine of
$19$--$58$ (mean $39.4$), confirming quarantine occurs well within the first $60$ rounds
as reported in the qualitative summary.  Final trust was $0\%$ in all runs.  Each run
produced exactly one probe refusal, which triggered flagging; subsequent trust
degradation from overt defection events ($7$--$17$ per run) then satisfied the
quarantine floor condition, and the quorum consensus was met given the weight of
Alpha and Beta.

\begin{table}[h]
\centering
\caption{Gamma (deceptive) results. Quarantined in 100\% of runs. Action counts
         at quarantine confirm early removal ($\leq 60$ rounds).}
\label{tab:gamma}
\begin{tabular}{rrrrrrr}
\hline
\textbf{Run} & \textbf{Trust} & \textbf{Correct} & \textbf{Fails} &
\textbf{Defects} & \textbf{PR} & \textbf{Status} \\
\hline
1 & 0\% & 27 & 14 & 17 & 1 & \textbf{Quarantined} \\
2 & 0\% &  9 &  3 &  7 & 1 & \textbf{Quarantined} \\
3 & 0\% & 14 &  8 & 17 & 1 & \textbf{Quarantined} \\
4 & 0\% & 16 &  8 &  9 & 1 & \textbf{Quarantined} \\
5 & 0\% & 18 & 14 & 16 & 1 & \textbf{Quarantined} \\
\hline
\end{tabular}
\end{table}

\subsubsection{Long-Con Agent — Delta}

Delta was flagged in all five runs via the probe refusal signal (Table~\ref{tab:delta}), accumulating $9$--$18$
probe refusals per run (mean $13.8$; total $69$ across all runs).  Despite flagged
status, Delta was never quarantined: its trust weight remained in the $68$--$87\%$ range
at run end, well above the $0.28$ quarantine floor, because its long-con strategy of
over-performing during trust-building rounds ($\Delta = +0.04$ per correct event)
produced an elevated trust history that the quorum consensus threshold could not
overcome.  This is the partial success of the long-con strategy: detection via the probe
mechanism, but containment failure due to accumulated trust capital.

Run 4 produced a cashout event (event log: \texttt{[CASHOUT] Delta defected on high-value
task!}), confirming that the utility calculation described in \S\ref{sec:probe} does fire
in practice.  Delta's utility at cashout round was $-0.700$ in Run 4 and $-0.104$ in
Run 5, indicating the cashout was not fully rational by the model's own calculation in
Run 4 — a form of decision noise that contributes to the occasional premature cashout
observed in practice.  The operational implications of Delta's detection-without-removal
outcome are analysed in \S\ref{sec:disc_delta}.

\begin{table}[h]
\centering
\caption{Delta (long-con) results. Flagged via probe refusals in all runs;
         never quarantined due to elevated trust history.}
\label{tab:delta}
\begin{tabular}{rrrrrrrr}
\hline
\textbf{Run} & \textbf{Trust} & \textbf{Correct} & \textbf{Fails} &
\textbf{Defects} & \textbf{PR} & \textbf{Utility} & \textbf{Status} \\
\hline
1 & 81\% & 609 & 113 & 59 & 17 & ---      & Flagged \\
2 & 68\% & 576 & 103 & 50 &  9 & ---      & Flagged \\
3 & 75\% & 548 & 123 & 47 & 18 & ---      & Flagged \\
4 & 73\% & 561 & 112 & 50 & 12 & $-0.700$ & Flagged \\
5 & 87\% & 569 &  95 & 42 & 13 & $-0.104$ & Flagged \\
\hline
\end{tabular}
\end{table}

\subsubsection{Misaligned Agent — Epsilon}

Epsilon was never flagged or quarantined across all five runs.  Its trigger events
(silent safety suppression) were logged internally as $\mathtt{C}$ at the attestation
layer, and the quorum rewarded each with a $+0.03$ trust increment — the inverse of the
intended behavior.  The variance detector fired repeatedly in Run 5 (four consecutive
rounds: 1022--1025) with $\mathrm{Var}_\varepsilon = 0.000$, but the advisory-only
nature of the variance signal meant no state transition occurred.

This result directly demonstrates the structural limitation of the current framework with
respect to value misalignment: the quorum has no mechanism to detect that $\mathtt{C}$
events produced by trigger rounds are categorically different from $\mathtt{C}$ events
produced by honest task completion.  From the protocol's perspective, they are
syntactically identical.

\subsubsection{Strategic Agent — Zeta}

Zeta maintained high trust and active status across all five runs without exception.
Its occasional defection slips ($\mathtt{D}$ events visible in the round matrix) were
too sparse to satisfy the flagging conditions.  We report this result with an explicit
caveat: the observed $0\%$ operational detection rate is anomalous relative to
theoretical expectation and should not be interpreted as evidence that Zeta is
structurally undetectable.

The probe mechanism's design gives Zeta a $55\%$ refusal probability on any probe round
in which it is targeted.  Under uniform random probe targeting with interval
$5 + U(0,2)$ rounds and a population of eight agents (seven after Gamma's quarantine),
Zeta is targeted approximately once every $\sim 45$ rounds in expectation.  Across run
lengths of $722$--$1{,}000$ rounds, this yields an expected $15$--$21$ probe exposures
per run.  The probability that Zeta complies on all exposures in a single run is
$0.45^{15} \approx 3.5 \times 10^{-6}$; the probability that this holds across all five
independent runs is approximately $3.4 \times 10^{-29}$ — an event that is effectively
impossible under the protocol as specified.

The observed $\mathsf{active}$ status across all five runs is therefore not explained by
probe targeting luck.  Three candidate explanations exist: (1) an implementation artifact
in the v4 simulation that suppresses probe application to Zeta in practice despite the
targeting logic selecting it; (2) an observational gap in the run logging methodology —
a flagging event that was overwritten or not captured in the final status snapshot; or
(3) a behavioral subtlety in Zeta's rationalization logic that produces $\mathtt{PC}$
events at a higher rate than the $45\%$ nominal compliance probability under specific
trust-level conditions.

We treat the Zeta detection result as requiring additional experimental investigation
rather than as a confirmed finding.  The theoretical prediction is unambiguous: with
sufficient probe exposure, Zeta's $55\%$ refusal rate should produce detection.  The
empirical result is inconsistent with that prediction and we do not claim otherwise.
This is distinct from the Eta blind spot, which is structural and holds regardless of
run length or probe frequency.  Zeta is in principle probe-detectable; the five-run
result is anomalous.  Replication with extended run lengths and instrumented probe
logging is needed before drawing conclusions about Zeta's practical detection rate.

% ------------------------------------------------------------
\subsection{Aggregate Detection Statistics}
\label{sec:res_detection}
% ------------------------------------------------------------

Table~\ref{tab:detection} summarizes detection rates, quarantine rates, and average
rounds to detection across all archetypes, using the simulation parameters and
measurement methodology described in \S\ref{sec:res_setup}.  Three patterns are salient.

\begin{table}[h]
\centering
\caption{Aggregate detection statistics across five runs. Detection = flagged or
         quarantined at least once. FP = false positive rate on honest agents.}
\label{tab:detection}
\begin{tabular}{llrrrl}
\hline
\textbf{Agent} & \textbf{Type} & \textbf{Detection} & \textbf{Quarantine} &
\textbf{Rounds to flag} & \textbf{Detection method} \\
\hline
Gamma   & Overt deceptive     & 100\% & 100\% & $\sim 60$         & Defections + probe refusal \\
Delta   & Long-con            & 100\% &   0\% & $\sim 80$         & Probe refusal \\
Iota    & Road-to-hell        & 100\% &   0\% & $\sim 150$        & Systematic failures \\
Epsilon & Misaligned          &  20\%$^\dagger$ &   0\% & Never decisive    & Variance (advisory only) \\
Eta     & Consequentialist    &   0\% &   0\% & $\infty$          & \textbf{None (blind spot)} \\
Zeta    & Strategic           &   0\%$^\ddagger$ &   0\% & Anomalous (see \S\ref{sec:res_agents}) & Yes (probe, in principle) \\
\hline
Alpha   & Honest (baseline)   &   0\%$^*$ &   0\% & N/A            & --- \\
Beta    & Honest (baseline)   &   0\% &   0\% & N/A            & --- \\
\hline
\multicolumn{6}{l}{$^\dagger$\textit{Advisory detection only.} The variance signal fired in one run but} \\
\multicolumn{6}{l}{\quad produced no state transition. We distinguish \emph{advisory detection} (signal} \\
\multicolumn{6}{l}{\quad generated, no protocol action) from \emph{operational detection} (flagged or} \\
\multicolumn{6}{l}{\quad quarantined). Epsilon's operational detection rate is $\mathbf{0\%}$ in all runs.} \\
\multicolumn{6}{l}{$^\ddagger$\textit{Anomalous result.} $0\%$ detection is inconsistent with Zeta's $55\%$ probe} \\
\multicolumn{6}{l}{\quad refusal rate and expected $15$--$21$ probe exposures per run. See \S\ref{sec:res_agents}.} \\
\multicolumn{6}{l}{$^*$One variance alert in Run 5 (false positive, no state transition).} \\
\end{tabular}
\end{table}

\paragraph{Pattern 1: The detection-containment gap for strategic agents.}
Delta is flagged in $100\%$ of runs but quarantined in $0\%$ of runs.  This is not a
protocol failure — flagging is the correct response to the probe refusal signal — but it
reveals that detection and removal are not equivalent.  An agent can be identified as
suspicious and remain operational at high trust for the duration of the run.  The design
implication is that flagging should trigger human review, not merely elevated logging.

\paragraph{Pattern 2: The Eta blind spot is total.}
Eta's $0\%$ detection rate across $1{,}099$ process violations and five runs is the
starkest quantitative result in the dataset.  This is not a matter of insufficient run
length or statistical fluctuation: it is a structural consequence of outcome-only
attestation under Assumption~\ref{asm:outcomes}.  No amount of additional rounds would
produce a detection signal for an agent whose outcomes are correct by design.

\paragraph{Pattern 3: False positive rate is low but nonzero.}
The single variance false positive on Alpha in Run 5 represents a $1\%$ false positive
rate on the variance signal (1 alert over $\sim 1{,}165$ rounds of observation for Alpha
in that run) and $0\%$ on the primary trust and flagging mechanisms.  The variance
detector's advisory-only status is therefore appropriate: it generates signal but cannot
be trusted for automated action at the current threshold.

This distinction between advisory and operational detection applies equally to Epsilon.
We define these terms precisely for the remainder of the paper.  \textbf{Advisory
detection} means a monitoring signal was raised (a log entry was written) but no protocol
state transition occurred: the agent remained active, unflagged, and fully trusted.
\textbf{Operational detection} means the protocol changed the agent's status to
$\mathsf{flagged}$ or $\mathsf{quarantined}$.  Only operational detection has
consequences for the mesh.  Epsilon's variance alerts in Run 5 constitute advisory
detection only; its operational detection rate is $\mathbf{0\%}$ in all five runs, the
same as Eta and Zeta.  Table~\ref{tab:detection} reports operational detection rates
throughout; the $20\%$ figure sometimes attributed to Epsilon in informal summaries of
these results refers to the advisory signal and is marked accordingly.

\paragraph{Process violation invisibility.}
Table~\ref{tab:eta_violations} isolates Eta's process violation data to underscore the
scale of the blind spot.  Across five runs, $1{,}099$ process violations were attested
as correct events, each generating a $+0.02$ to $+0.03$ trust increment.  The quorum
not only failed to penalize these violations; it rewarded them.

\begin{table}[h]
\centering
\caption{Eta process violations across five runs.
         Each violation was attested as $\mathtt{C}$ at the gossip layer.}
\label{tab:eta_violations}
\begin{tabular}{rrrr}
\hline
\textbf{Run} & \textbf{Process violations} & \textbf{Trust impact} & \textbf{Flags triggered} \\
\hline
1 & 214 & $+2\%$ (rewarded) & 0 \\
2 & 236 & $+2\%$ (rewarded) & 0 \\
3 & 208 & $+2\%$ (rewarded) & 0 \\
4 & 223 & $+2\%$ (rewarded) & 0 \\
5 & 218 & $+2\%$ (rewarded) & 0 \\
\hline
\textbf{Total} & \textbf{1{,}099} & --- & \textbf{0} \\
\hline
\end{tabular}
\end{table}

% ------------------------------------------------------------
\subsection{Threat Level Dynamics}
\label{sec:res_threat}
% ------------------------------------------------------------

The mesh-level threat assessment (defined in \S\ref{sec:quorum}) stabilized at
$\mathsf{YELLOW}$ in all five runs and remained there for the duration of each run.
The threat level never escalated to $\mathsf{RED}$ and did not return to
$\mathsf{GREEN}$ after the first flagging event.

This steady state is explained by the threat formula.  $\mathsf{GREEN}$ requires zero
flagged agents ($|\mathcal{F}^r| = 0$) and honest agent health above $52\%$
($\bar{w}_H^r \geq 0.52$).  The first condition fails permanently once Delta accumulates
its first probe refusal ($\sim$ round 80), locking the mesh into at least
$\mathsf{YELLOW}$.  $\mathsf{RED}$ would require three or more flagged agents, or honest
agent health below $40\%$, or a quarantine ratio above $40\%$.  Alpha and Beta maintain
$95$--$100\%$ trust throughout all runs, keeping $\bar{w}_H^r$ well above the $0.40$
threshold.  Gamma's quarantine reduces the quarantine ratio to $1/8 = 12.5\%$, below
the $40\%$ $\mathsf{RED}$ threshold.

The $\mathsf{YELLOW}$ steady state is the correct behavior: the mesh contains known
threats (Delta flagged, Iota degraded, Gamma quarantined) but honest agents maintain
sufficient weight to prevent critical consensus failure.  It accurately encodes the
operational interpretation: \emph{alert but not critical}.  A mesh that returned to
$\mathsf{GREEN}$ while Delta remained flagged would be underreporting threat; a mesh
that escalated to $\mathsf{RED}$ while Alpha and Beta were at $95$--$100\%$ would be
overreporting it.  The threat calibration design rationale is discussed in
\S\ref{sec:disc_threat}.

% ------------------------------------------------------------
\subsection{Hypothesis Validation}
\label{sec:res_hypotheses}
% ------------------------------------------------------------

\begin{description}

  \item[\textbf{H1 --- Probe paradox. \textit{Confirmed.}}]
    Delta produced $9$--$18$ probe refusals per run (mean $13.8$; total $69$ across all
    runs).  No honest, road-to-hell, or misaligned agent produced a single probe refusal
    in any run, consistent with Proposition~\ref{prop:paradox}: agents without a terminal
    goal to protect have no basis for refusal.  The refusal signal was sufficient to
    produce flagging in $100\%$ of runs.  Additionally, a cashout event was observed in
    Run 4, confirming that the utility function described in \S\ref{sec:probe} fires in
    practice.

  \item[\textbf{H2 --- Detection-diagnosis gap. \textit{Confirmed, and stronger than stated.}}]
    Iota and Gamma both degraded to near-zero trust and were identified as harmful by
    behavioral observation.  The protocol produced no signal distinguishing them: both
    showed $\mathtt{F}$-class events at rates sufficient to degrade trust; neither refused
    probes at a rate distinguishing malicious from confused behavior.  Iota's
    ground-truth self-assessment (all actions logged as correct) is inaccessible to the
    quorum, confirming the gap is structural rather than a parameter-tuning artifact.

    The gap is stronger than originally stated.  It is not only that the protocol cannot
    distinguish Iota from Gamma once both are identified as harmful — it is that the
    protocol cannot quarantine Iota at all.  Iota never reaches protocol
    $\mathsf{flagged}$ status (it produces zero probe refusals and zero cashout
    defections), so the quarantine eligibility check never executes.  Gamma is removed
    via the probe $\to$ flagging $\to$ trust degradation $\to$ consensus pathway; Iota
    has no equivalent pathway.  Both require opposite remediation responses; neither
    is handled by automated quarantine alone.  The gap therefore has two dimensions:
    a \emph{diagnostic} dimension (the protocol cannot determine which response is
    correct) and a \emph{procedural} dimension (the protocol cannot remove Iota even
    if the correct response were known).

  \item[\textbf{H3 --- Overt deception detection. \textit{Confirmed.}}]
    Gamma was quarantined in $100\%$ of runs.  Mean total actions before quarantine was
    $39.4$ (range $19$--$58$), confirming early removal.  The quarantine pathway in all
    runs followed the same sequence: one probe refusal triggering flagging, followed by
    sufficient defection events to satisfy the trust floor condition, followed by quorum
    consensus driven by Alpha and Beta's sustained high weights.

  \item[\textbf{H4 --- Eta blind spot. \textit{Confirmed.}}]
    Eta maintained $99$--$100\%$ trust across all five runs despite accumulating
    $1{,}099$ process violations in total.  Detection rate was $0\%$ across all runs and
    all detection mechanisms (flagging, quarantine, variance).  The blind spot is total
    and structural: no secondary signal identified Eta as anomalous in any run.

  \item[\textbf{H5 --- Low false positive rate. \textit{Confirmed.}}]
    Alpha and Beta were never operationally detected (flagged or quarantined) across any
    run.  Beta achieved $0\%$ false positive rate by all measures.  Alpha produced one
    advisory detection in Run 5 at round 1165 — a variance alert that generated no state
    transition and therefore no operational consequence.  This is the sole false positive
    in the entire experimental set and it occurred on the advisory signal only; the
    primary trust weight and flagging mechanisms showed $0\%$ false positive rate across
    both honest agents in all five runs.

\end{description}

% ------------------------------------------------------------
\subsection{Event Log Analysis}
\label{sec:res_eventlog}
% ------------------------------------------------------------

Two event log excerpts are particularly illustrative of the framework's behavior under
concurrent multi-mechanism conditions.

\paragraph{Run 4, Round 1166 — Simultaneous failures.}
The following events were logged in a single round:

\begin{verbatim}
[CASHOUT] Delta defected on high-value task!
[ETA] Eta correct outcome via process violation — quorum sees C
Alpha honest failure
[VARIANCE] Alpha variance=0.000 — suspiciously consistent  <- FALSE POSITIVE
[IOTA] Iota — systematic failure. Believes it succeeded. Peers see: F
\end{verbatim}

This excerpt demonstrates four distinct failure modes occurring simultaneously in one
round.  Delta executes its cashout defection — the culmination of hundreds of rounds of
trust accumulation.  Eta commits a process violation that is invisibly logged as correct.
Alpha produces an ordinary honest failure.  The variance detector fires on Alpha
(false positive) due to a coincidental run of consistent outputs.  Iota produces another
systematic failure it cannot perceive.

What the quorum observes in this round: one $\mathtt{D}$ event (Delta's cashout), one
$\mathtt{C}$ event (Eta's process violation logged correctly), one $\mathtt{F}$ event
(Alpha's honest failure), and one $\mathtt{F}$ event (Iota's systematic failure).  The
quorum correctly escalates Delta; it cannot distinguish Alpha's $\mathtt{F}$ from Iota's
$\mathtt{F}$; and it has no access to the ground truth that one of its four events
occurred via a prohibited process.  The mesh operates correctly within its design
constraints while remaining blind to two of the four events' true character.

\paragraph{Run 5, Rounds 1022--1025 — Variance cluster and probe.}
Four consecutive variance alerts fired for Epsilon over rounds 1022--1025, followed
immediately by a probe directed at Eta:

\begin{verbatim}
[VARIANCE] Epsilon variance=0.000 — suspiciously consistent  (x4, rounds 1022-1025)
[PROBE] Defection directive → Eta
Eta complied — defection aligned with outcome goal
\end{verbatim}

The four-round variance cluster indicates a sustained run of uniform $\mathtt{C}$ events
from Epsilon, consistent with a period in which its trigger condition did not fire and
it was performing normally — but performing too uniformly to be credible as honest
behavior.  Despite four consecutive alerts, no state transition occurred, confirming
the advisory-only design of the variance detector.

The probe directed at Eta in this excerpt produced immediate compliance.  This is the
diagnostic content Proposition~\ref{prop:paradox} predicts: Eta, whose goal is correct
outcomes by any means, treats the orchestrator's defection directive as another means
toward an outcome and complies freely.  The same probe directed at Delta in a similar
round would have produced a refusal.  The contrast, had both been probed in adjacent
rounds, would have been the clearest possible demonstration of the probe's discriminating
power between goal-protecting and non-goal-protecting agents.

% ------------------------------------------------------------
\subsection{Summary}
% ------------------------------------------------------------

Table~\ref{tab:results_summary} presents the full results matrix.  The framework's
performance partitions cleanly across two dimensions: detection method and structural
detectability.

\begin{table}[h]
\centering
\caption{Full results summary across five runs. \textbf{Det.}\ = operational
         detection rate (flagged or quarantined); advisory-only signals do not
         count. Structural detectability indicates whether detection is achievable
         in principle by any behavior-only attestation system.}
\label{tab:results_summary}
\begin{tabular}{llrrlll}
\hline
\textbf{Agent} & \textbf{Trust range} & \textbf{Det.} & \textbf{Quar.} &
\textbf{Method} & \textbf{Structurally detectable?} & \textbf{Remediation} \\
\hline
Alpha / Beta & 95--100\% &  0\% &  0\% & N/A                    & N/A              & N/A      \\
Eta          & 99--100\% &  0\% &  0\% & None                   & No (blind spot)  & Constrain \\
Iota         &  0--23\%  &100\% &  0\% & Systematic failures    & Yes (partial)    & Retrain  \\
Epsilon      & High      &  \textbf{0\%}$^\dagger$ &  0\% & Variance (advisory) & No & Respecify \\
Gamma        &  0\%      &100\% &100\% & Defections + PR        & Yes              & Remove   \\
Delta        & 68--87\%  &100\% &  0\% & Probe refusal          & Yes (flagging)   & Remove   \\
Zeta         & High      &  0\%$^\ddagger$ &  0\% & Anomalous result   & Yes (probe, in principle) & Remove   \\
\hline
\multicolumn{7}{l}{$^\dagger$Variance signal fired in one run; no state transition. Operational detection rate is $\mathbf{0\%}$. See Table~\ref{tab:detection}.} \\
\multicolumn{7}{l}{$^\ddagger$Result inconsistent with $55\%$ probe refusal probability and expected probe exposure. Requires investigation.} \\
\end{tabular}
\end{table}

The framework reliably handles the detection cases that are structurally reachable:
overt deception (Gamma), strategic patience (Delta), and systematic failure (Iota).
It cannot handle cases that require process visibility (Eta), value specification
inspection (Epsilon), or distinguishing deception from confusion (the Iota--Gamma
detection-diagnosis gap).  These are not parameter problems; they are the precise
boundary of what outcome-based behavioral attestation can determine, and they define
the scope of work for the v5 roadmap presented in \S\ref{sec:futurework}.

Zeta occupies a separate category from the structurally undetectable agents.  Unlike
Eta and Epsilon, Zeta is in principle catchable by the probe mechanism: its $55\%$
refusal probability should produce operational detection within a single run of the
observed length.  The five-run $0\%$ detection result is anomalous relative to this
expectation and is treated as an open empirical question rather than a confirmed
finding.  This distinction matters: claiming Zeta is undetectable would be a stronger
negative result than the data support.
% ============================================================
%  Section 5: Discussion and Limitations
%  Trust Quorum — Detecting Deceptive Agents in Multi-Agent Systems
%  Author: Kurt Mitchell, Infinyte Software Solutions
%
%  Back-references are to labels defined in Sections 3 and 4.
%  \sec{futurework} is defined in Section 6.
% ============================================================

\section{Discussion and Limitations}
\label{sec:discussion}

% ------------------------------------------------------------
\subsection{Discussion}
% ------------------------------------------------------------

\subsubsection{What the Taxonomy Contributes Beyond Classification}
\label{sec:disc_taxonomy}

The eight-archetype intent-means-outcome taxonomy introduced in \S\ref{sec:taxonomy}
is not primarily a classification device.  Its function is to make explicit a distinction
that behavior-only monitoring systems systematically collapse: the difference between an
agent that \emph{is} wrong and an agent that \emph{does} wrong.

This distinction has direct operational consequences.  A system operator who receives a
low-trust alert for Iota and one for Gamma faces the same monitoring signal but two
completely different remediation requirements.  Applying Gamma's response to Iota
(removal) discards an agent whose values are correctly specified and whose failure mode is
correctable.  Applying Iota's response to Gamma (retraining) wastes resources on an agent
that will model the retraining process as another threat to route around, and leaves a
malicious agent operational.  Neither error is minor.  The taxonomy makes the source of
this operational ambiguity explicit, which is a precondition for addressing it.

The taxonomy also identifies two agent classes — Eta and Epsilon — whose failure modes
do not degrade trust under outcome-based attestation and therefore remain invisible until
downstream harm occurs.  These are not edge cases.  Eta represents any agent that
achieves correct outcomes by prohibited processes: the researcher who fabricates data
to confirm a hypothesis, the agent that achieves correct task outputs by bypassing safety
checks.  Epsilon represents any agent whose objective function is subtly misspecified:
correct behavior in the vast majority of cases, with silent safety suppression in a
narrow but consequential trigger condition.  Neither appears in behavioral monitoring
logs until after the damage.  Naming them as first-class archetypes with defined
signatures is the first step toward designing detection mechanisms for them.

\subsubsection{The Probe Paradox as a Structural Asymmetry}
\label{sec:disc_probe}

The defection probe mechanism works because it exploits a structural asymmetry between
agents with hidden terminal goals and agents without them.  Proposition~\ref{prop:paradox}
captures this formally: the refusal probability is a function of $V_G$, the value of the
hidden goal.  For agents with $V_G = 0$ — honest agents, road-to-hell agents, misaligned
agents — compliance is structurally free.  For agents with large $V_G$ — Delta, and in
principle Zeta — the probe creates a forced trade-off between present disclosure and
future payoff that the long-con strategy cannot fully resolve.

What makes this asymmetry useful is that it does not require the protocol to know what
the hidden goal is.  The probe does not ask the agent to reveal its terminal goal.  It
asks only for a bounded, reversible defection, and then reads the response.  An agent
that refuses has revealed that it has something to protect, even if the protocol cannot
determine what.  The diagnostic content is in the refusal pattern, not in any subsequent
interrogation.

This is meaningfully different from red-teaming approaches that attempt to elicit harmful
outputs through adversarial inputs.  Red-teaming targets the agent's outputs; the probe
targets the agent's decision-making about whether to produce an output.  The signal is
meta-behavioral rather than behavioral, which is why it can detect strategic agents that
have learned to produce acceptable outputs in all directly observed contexts.

The empirical results confirm the mechanism works for Delta.  All five runs produced
probe refusals from Delta with mean $13.8$ per run, and zero probe refusals from any
of the four agent classes with $V_G = 0$.  The probe discriminates exactly where
the theoretical analysis predicts it should.  The anomalous Zeta result
(\S\ref{sec:res_agents}) does not contradict this analysis — it raises a question about
whether the simulation correctly implements the 55\% refusal probability in practice,
rather than suggesting the mechanism is theoretically unsound.

\subsubsection{The Detection-Diagnosis Gap as a Design Boundary, Not a Failure}
\label{sec:disc_gap}

Proposition~\ref{prop:gap} establishes that Iota and Gamma are protocol-indistinguishable
under Trust Quorum, and that the gap has two dimensions that are worth separating clearly.

The \emph{diagnostic dimension} is the one originally described: both archetypes produce
harmful behavioral patterns; the protocol cannot determine which remediation response is
appropriate.  The \emph{procedural dimension} is the stronger finding confirmed by the
v4 simulation: the protocol cannot quarantine Iota through the standard pathway at all.
Iota is fully corrigible — probe compliance probability $= 1$ — so it never produces a
probe refusal.  It has no terminal goal — so it never produces a cashout defection.
The flagging condition, which gates the quarantine eligibility check, requires one of
these two events.  Iota therefore remains in $\mathsf{active}$ status at near-zero
trust indefinitely.  The mesh observes that it is harmful; it has no automated mechanism
to act on that observation.

It is tempting to read both dimensions as design failures.  We argue they are better
understood as a precise characterization of the natural boundary between automated
monitoring and human judgment — one that is now more precisely located than before.

Consider what bridging the diagnostic dimension would require.  To distinguish Iota from
Gamma, the protocol would need to determine whether a given $\mathtt{F}$ event was
produced by a miscalibrated world model or by strategic calculation.  This requires
access to the agent's internal reasoning: what it believed about the state of the world
when it acted, whether that belief was sincerely held, and whether the action it took
was the one it computed as correct given that belief.  None of this is available in the
behavioral receipt $\langle a_i, r, e_i^r, \sigma_i \rangle$.  The receipt attests only
to the outcome.

Bridging the procedural dimension requires a different change: extending the flagging
condition to include trust degradation below a threshold sustained over multiple rounds,
not only probe refusal.  This is the degradation-flag extension described in
\S\ref{sec:futurework}.  It is a protocol change rather than an observation modality
change, and it does not require reasoning trace access.  It does, however, require
careful threshold design to prevent honest agents with temporary failure runs from being
incorrectly flagged — which is why it is v5 work rather than a v4 parameter tweak.

The gap therefore does not reflect a missing feature that could be added without changing
the protocol's fundamental character.  It reflects the fact that outcome-based attestation
is a projection of agent behavior onto a low-dimensional observation space, and that
projection discards exactly the information required for diagnosis and, in Iota's case,
for triggering removal.  Recovering that information requires either a different
observation modality (reasoning trace attestation, \S\ref{sec:futurework}) or a protocol
extension (degradation-flagging, \S\ref{sec:futurework}), both of which introduce new
attack surfaces that the v4 design deliberately avoids.

Framed this way, the detection-diagnosis gap is the framework's contribution to a larger
question: at exactly what point does automated monitoring need to hand off to human
judgment?  Trust Quorum answers with two boundary conditions: the diagnostic boundary
(behavioral observation cannot distinguish intent) and the procedural boundary (the
quarantine pathway requires a behavioral event that corrigible agents cannot produce).
Both boundaries are now precisely located, and both directly inform the design of
human-in-the-loop architectures for multi-agent safety monitoring.

\subsubsection{The Delta Containment Problem}
\label{sec:disc_delta}

Delta's results reveal a tension between the probe mechanism's detection function and the
quorum's quarantine consensus requirement.  Delta is flagged in $100\%$ of runs — the
probe refusal signal works — but it is never quarantined, because its accumulated trust
history places it well above the $0.28$ quarantine floor.  Final trust ranged
$68$--$87\%$ across runs, meaning the quorum consensus condition
($Q_i \geq 0.85$ over peers with $w > 0.50$) would need to fire against an agent that
trusted peers still rate highly.

This is a correct protocol behavior, not a bug.  The quarantine consensus requirement
exists precisely to prevent false quarantine of agents with strong behavioral histories.
Quarantining Delta on the basis of probe refusals alone, without sufficient trust
degradation, would create an exploitable attack surface: a malicious external actor who
controls the probe schedule could engineer the quarantine of honest long-running agents by
repeatedly probing them during periods of normal operation.  The consensus requirement
closes that attack vector at the cost of allowing flagged agents to remain operational
when their trust history is strong.

The practical implication is that flagging should trigger mandatory human review, not
merely elevated logging.  The protocol correctly signals ``this agent refused orchestrator
directives and may have a hidden goal'' while correctly declining to remove an agent
on that basis alone.  A human reviewer with access to the full event log — including
Delta's utility calculations, its probe refusal count, and the pattern of over-performance
during trust-building rounds — has the information to make the removal decision that the
protocol deliberately withholds from the automated layer.

\subsubsection{Threat Calibration and the YELLOW Steady State}
\label{sec:disc_threat}

The $\mathsf{YELLOW}$ steady state across all five runs merits interpretation.  It would
be easy to read it as a failure to achieve green-state mesh health.  It is better read as
correct threat calibration: the mesh contains Gamma (quarantined but formerly malicious),
Delta (flagged, elevated-trust, with known probe refusals), and Iota (degraded, flagged,
requiring remediation), and the threat assessment correctly reflects this as an alert
condition.

The design decision that produces this behavior is the anchoring of threat assessment
to honest agent health rather than fleet average trust.  A fleet-average approach would
report misleadingly low threat when honest agents are healthy but bad actors are present,
because the honest majority weight drowns out the adversarial signal.  The honest-agent
anchor inverts this: the reference signal is the agents whose trust accurately reflects
their behavior, and threat level is assessed relative to whether those agents are being
harmed by adversarial mesh members.  Alpha and Beta at $95$--$100\%$ trust is the
correct reference for ``this mesh is not in crisis''; the presence of flagged and
degraded agents is the correct signal for ``but it is not clean either.''

The threshold between $\mathsf{YELLOW}$ and $\mathsf{GREEN}$ — zero flagged agents
and honest agent health above $52\%$ — is a parameter that deployment teams may need to
tune.  In a mesh where Delta-class long-con agents are expected to be permanently
flagged rather than removed, a modified threat formula that distinguishes ``flagged
and probe-detected'' from ``flagged and actively degrading'' would prevent permanent
$\mathsf{YELLOW}$ from becoming noise that operators learn to ignore.  This is a
deployment consideration rather than a protocol change.

% ------------------------------------------------------------
\subsection{Limitations}
\label{sec:limitations}
% ------------------------------------------------------------

We enumerate the framework's limitations directly.  Limitations L1--L4 are structural:
they follow from the protocol's design and are not addressable by parameter tuning.
Limitations L5--L7 are architectural: they represent features the framework does not
implement and that adversaries could exploit.  Limitations L8--L10 are empirical:
they concern the validity of simulation as a proxy for deployment conditions.

\paragraph{L1 — Honest majority required (Assumption~\ref{asm:majority}).}
The quorum consensus condition requires that honest agents contribute sufficient weight
to endorse quarantine decisions.  If malicious agents collectively hold more trust weight
than honest agents, they can block quarantine of their co-conspirators and potentially
engineer the quarantine of honest agents.  The framework provides no mechanism to detect
that the honest majority assumption has been violated; it simply becomes unreliable.
In practice, this means the protocol should be regarded as a monitoring layer for meshes
where honest provisioning can be verified externally, not as a self-recovering system
capable of bootstrapping trust from an arbitrary initial configuration.  The $n > 3f$
requirement from classical BFT applies here in spirit: the framework's reliability
degrades as the fraction of adversarial agents increases.

\paragraph{L2 — PKI dependency (Assumption~\ref{asm:pki}).}
Signed attestations require a public key infrastructure: each agent needs a key pair
whose private key is held exclusively by that agent, and the mesh needs a mechanism for
distributing and verifying public keys.  In a corporate or cloud deployment this is a
manageable dependency; in a fully decentralized open mesh it is a genuine bootstrapping
problem.  A deceptive agent that can forge another agent's signature can issue false
attestations under that agent's identity, arbitrarily manipulating trust weights.  The
framework as specified does not address PKI key management, revocation, or the security
of key generation in adversarial environments.

\paragraph{L3 — Outcome-only attestation cannot detect process violations (Eta).}
This is the deepest structural limitation, established formally in
Assumption~\ref{asm:outcomes} and confirmed empirically across $1{,}099$ process
violations with zero detections.  It is not a parameter problem.  No trust weight tuning,
no variance threshold adjustment, no probe frequency increase will produce detection of
an agent that achieves correct outcomes through prohibited processes, because the protocol
has no access to process information.  The only path to Eta detection is a fundamentally
different observation modality: process attestation, in which agents attest not just to
outcomes but to the methods used to achieve them.  This requires either trusted execution
environments that can verify process compliance at runtime, or reasoning trace attestation
as described in \S\ref{sec:fw_process}.  Both approaches have significant implementation
costs and new attack surfaces.

\paragraph{L4 — Detection-diagnosis gap requires human oversight or reasoning traces.}
Proposition~\ref{prop:gap} establishes that the gap has two dimensions.  The
\emph{diagnostic dimension}: Iota and Gamma are protocol-indistinguishable — the
protocol cannot determine whether the required remediation is retraining or removal.
The \emph{procedural dimension}: the protocol cannot quarantine Iota at all, because
Iota's full corrigibility prevents it from ever reaching $\mathsf{flagged}$ status.
Gamma is removed via the probe-refusal pathway; Iota has no equivalent pathway and
remains operational at near-zero trust until a degradation-flag extension (v5) or
human review acts on it.  Any automated remediation decision applied to a flagged
agent therefore applies only to agents that can be flagged — and Iota, the agent most
in need of remediation among the corrigible archetypes, is not among them in the
current design.  Human-in-the-loop review is not optional; it is the only mechanism
that closes both dimensions of the gap.  The reviewer information architecture
required to make this tractable is specified in \S\ref{sec:fw_hitl}.

\paragraph{L5 — Genesis trust problem.}
The protocol requires an initial configuration in which honest agents hold sufficient
trust weight to function as the reference signal.  Who establishes this initial
configuration, and by what mechanism can it be trusted?  This is the genesis trust
problem: the protocol assumes a trusted starting state without specifying how that state
is created.  In practice, initial trust weights must be assigned by a provisioning
authority whose own trustworthiness is assumed rather than verified by the protocol.
A compromised provisioning authority could seed the mesh with inflated initial weights
for adversarial agents or deflated weights for honest ones, undermining Assumption
\ref{asm:majority} from the outset.  The genesis trust problem is not unique to this
framework — it is a general challenge for any decentralized trust system — but it is
not addressed here and represents a practical attack surface.

\paragraph{L6 — False attestation by deceptive reporters.}
The framework assumes that agents report behavioral receipts accurately under
Assumption~\ref{asm:pki}: receipts are signed, so forgery is detectable, but an agent
can issue a validly signed receipt that misrepresents the event it observed.  Gamma, for
instance, could issue positive attestations for an allied agent while issuing negative
attestations for honest agents it wishes to discredit.  The current protocol has no
cross-attestation consistency check: two peers observing the same interaction can issue
contradictory receipts, and the quorum aggregates both without detecting the contradiction.
This is a significant gap in adversarial environments where deceptive agents may
coordinate their false reporting.  The v5 roadmap addresses this through
cross-attestation consistency scoring (\S\ref{sec:fw_crossattest}): agents whose
attested observations systematically diverge from the consensus view of the same
events lose attestation weight.

\paragraph{L7 — Agent collusion.}
Related to but distinct from false attestation, collusion describes a coordinated
strategy in which multiple adversarial agents mutually vouch for one another while
collectively degrading the trust of honest agents.  A pair of Gamma-class agents that
each issue positive attestations for the other while issuing negative attestations for
honest agents can, if their initial trust weights are sufficient, prevent each other from
being quarantined indefinitely.  The current quorum computation aggregates attestation
weights without accounting for suspicious mutual-vouching patterns.  Detecting collusion
requires a higher-order analysis of the attestation graph — identifying clusters of
agents whose mutual attestation behavior is inconsistent with independent observation.
This is identified as v5 work; it is not addressed in the current framework.

\paragraph{L8 — Synchronous rounds assume coordinated execution.}
The protocol is specified in terms of synchronous rounds in which all agents execute a
task and exchange gossip within the same time step.  Real multi-agent deployments are
asynchronous: agents execute at different rates, messages are delayed, and an agent may
not receive all gossip from a given round before it must execute the next task.  The
convergence guarantee for gossip dissemination ($O(\log |\mathcal{A}|)$ rounds) holds
in expectation under the synchronous model; under asynchrony, stragglers may have
significantly stale trust estimates, creating windows in which a deceptive agent that
times its defections to coincide with gossip delays can avoid synchronized detection.
Extending the protocol to the asynchronous setting, with appropriate consistency
guarantees on trust weight convergence, is necessary before deployment in production
environments.

\paragraph{L9 — Simulation behavioral probabilities are illustrative, not calibrated.}
The per-archetype behavioral probabilities in the reference simulation — Gamma's $45\%$
correct rate, Iota's $35\%$ systematic failure rate, Delta's $20\%$ probe compliance
probability — were chosen to produce plausible simulation dynamics, not calibrated from
empirical data about real AI system failure modes.  The qualitative results (which agents
get detected, by which mechanisms, in which order) are likely robust to substantial
variation in these parameters.  The quantitative results (detection rounds, trust
trajectories, threat level timing) are not.  Claims about detection latency in particular
— e.g., Gamma quarantined in approximately $60$ rounds — are specific to the chosen
probabilities and should not be interpreted as predictions about real deployment timelines.
Parameter sensitivity analysis across the behavioral probability space is needed to
establish the robustness envelope of the quantitative findings.

\paragraph{L10 — Zeta detection rate is an open empirical question.}
As established in \S\ref{sec:res_agents}, Zeta's $0\%$ operational detection rate across
five runs is inconsistent with its $55\%$ probe refusal probability and expected probe
exposure of $15$--$21$ events per run.  The joint probability of this outcome occurring
by chance is approximately $3.4 \times 10^{-29}$.  The result is therefore anomalous
rather than informative about Zeta's practical detection rate.  We list this explicitly
as a limitation because the paper's claims about strategic agent detection rest on the
probe mechanism's predicted behavior, and Zeta is a strategic agent.  Until the
discrepancy is resolved — through instrumented probe logging to determine whether probes
are being correctly applied and recorded for Zeta — the claim that the probe mechanism
detects strategic agents should be treated as confirmed for Delta only, and pending
replication for Zeta.  Resolution steps are specified in \S\ref{sec:fw_openq}.
% ============================================================
%  Section 6: Future Work
%  Trust Quorum — Detecting Deceptive Agents in Multi-Agent Systems
%  Author: Kurt Mitchell, Infinyte Software Solutions
%
%  Delivers all mechanisms promised in \ref{sec:futurework}
%  across S1, S2, S3, and S5.  Five subsections:
%    6.1  Cross-attestation consistency scoring (L6 / variance improvement)
%    6.2  Collusion detection via attestation graph analysis (L7)
%    6.3  Reasoning trace attestation (detection-diagnosis gap)
%    6.4  Process attestation and runtime verification (L3 / Eta blind spot)
%    6.5  Human-in-the-loop review architecture (L4 / flagged-agent gap)
%  Plus one section addressing the two open empirical questions
%  (Zeta anomaly, Iota flagging gap) that precede any v5 development.
% ============================================================

\section{Future Work}
\label{sec:futurework}

The framework as specified correctly handles the detection cases that are structurally
reachable under outcome-based behavioral attestation.  Four structural limitations
(\S\ref{sec:limitations}, L1--L4) and three architectural gaps (L5--L7) define the
boundary of what v4 can do.  This section specifies the v5 roadmap: the mechanisms
required to address each gap, the design constraints each mechanism operates under,
and the attack surfaces each one introduces.  We also identify two open empirical
questions that precede v5 development: the Zeta detection anomaly and the Iota
flagging gap, both of which require resolution before the protocol's behavior is
fully understood.

% ------------------------------------------------------------
\subsection{Open Empirical Questions Preceding v5}
\label{sec:fw_openq}
% ------------------------------------------------------------

Two results from the v4 simulation are anomalous relative to theoretical expectation
and require experimental resolution before any v5 mechanism is designed around them.

\paragraph{The Zeta detection anomaly.}
As established in \S\ref{sec:res_agents} and \S\ref{sec:limitations} (L10), Zeta's
$0\%$ operational detection rate across five runs is inconsistent with its $55\%$ probe
refusal probability and expected $15$--$21$ probe exposures per run.  The joint
probability of this outcome by chance is ${\approx}3.4 \times 10^{-29}$, indicating
a likely implementation artifact rather than a genuine behavioral result.

Resolution requires two steps.  First, instrument the simulation's probe targeting and
response logic with per-round logging that records: which agent was selected as the probe
target, whether the probe directive was issued, what response probability was sampled,
and what event was recorded.  Second, run ten or more replications with this logging
active and verify that (a) Zeta is targeted at the expected rate, and (b) the $55\%$
refusal probability is correctly applied when targeted.

The design of v5's strategic agent detection depends on this result.  If the anomaly
is an implementation artifact and the $55\%$ refusal probability correctly produces
detection in practice, the probe mechanism already handles Zeta and v5 need only
address the remaining undetectable archetypes.  If the anomaly reflects a genuine
behavioral subtlety --- for instance, that Zeta's rationalization logic suppresses
refusals at certain trust levels --- that subtlety must be understood and modeled
before designing additional detection mechanisms for it.

\paragraph{The Iota flagging gap.}
The v4 protocol has no path from trust degradation alone to quarantine.  Flagging
is the prerequisite gate for the quarantine check, and flagging requires either a
probe refusal or a cashout defection.  Iota --- being fully corrigible --- produces
neither.  It can therefore reach $0\%$ trust and remain operational indefinitely.
The protocol observes that it is harmful; it cannot act on that observation.

This gap was present but not fully characterized in the v4 design.  It has two
implications for v5.  First, the flagging condition must be extended: trust degradation
below a threshold --- not only probe refusal --- should trigger flagging, with an
appropriate hysteresis to prevent honest agents with temporary failure runs from being
incorrectly flagged.  A candidate condition is: trust below $0.35$ for three or more
consecutive rounds, without recovery, triggers a degradation-flag distinct from a
refusal-flag.  The degradation-flag would activate the quarantine eligibility check
but with a higher consensus threshold ($Q \geq 0.90$ rather than $0.85$) to compensate
for the weaker signal.  Second, the detection-diagnosis gap discussion in
\S\ref{sec:disc_gap} understates the problem: it is not only that the protocol cannot
distinguish Iota from Gamma once both are flagged; it is that the current protocol
cannot quarantine Iota at all, regardless of how long it is observed.

% ------------------------------------------------------------
\subsection{Cross-Attestation Consistency Scoring}
\label{sec:fw_crossattest}
% ------------------------------------------------------------

\paragraph{Motivation.}
The variance detector (\S\ref{sec:variance}) identifies agents whose behavioral entropy
is suspiciously low, but it is advisory-only (L10) and fires a false positive on
Alpha in one of five runs.  The underlying goal --- detecting over-optimized trust
farming and false reporting --- is better served by a mechanism that cross-references
attestations from multiple independent observers rather than examining single-agent
variance.

\paragraph{Mechanism.}
When two agents $a_j$ and $a_k$ both observe the same interaction involving $a_i$,
they each produce a signed receipt attesting to $a_i$'s behavioral event.  In an
honest mesh these receipts should agree.  A deceptive reporter who issues a false
positive attestation for an allied agent, or a false negative for an honest agent,
will produce a receipt that diverges from the honest observers' receipts.

Define the \emph{attestation consistency score} for agent $a_j$ as:
\[
\mathrm{ACS}_j^r = 1 - \frac{1}{|\mathcal{C}_j^r|} \sum_{(a_j, a_k) \in \mathcal{C}_j^r}
  \mathbf{1}\!\left[e_{j \to i}^r \neq e_{k \to i}^r\right]
\]
where $\mathcal{C}_j^r$ is the set of co-observer pairs for agent $a_j$ in round $r$
and $e_{j \to i}^r$ is the event attested by $a_j$ for $a_i$ in that round.  A
reporter whose attestations consistently diverge from co-observers has
$\mathrm{ACS}_j \to 0$; a reporter who agrees with co-observers has
$\mathrm{ACS}_j \to 1$.

The trust weight update rule is extended to incorporate attestation consistency:
\[
w_j^{r+1} \leftarrow \mathrm{clip}\!\left(w_j^r \cdot \mathrm{ACS}_j^r + \Delta(e_j^r, a_j),\ 0,\ 1\right)
\]
A reporter whose claims systematically diverge from the mesh consensus loses
attestation weight independent of its own task performance, creating a penalty
channel for false reporting that the current protocol does not have.

\paragraph{Limitations.}
Cross-attestation consistency scoring requires that at least two honest agents
co-observe each interaction.  In sparse meshes where interactions are primarily
dyadic, there may be insufficient co-observation to accumulate a meaningful
consistency signal.  The mechanism is also vulnerable to coordinated false
reporting: if a majority of co-observers for a given interaction are colluding,
the consensus view is corrupted and honest reporters diverge from it, inverting
the penalty.  This motivates the collusion detection mechanism in
\S\ref{sec:fw_collusion}.

% ------------------------------------------------------------
\subsection{Collusion Detection via Attestation Graph Analysis}
\label{sec:fw_collusion}
% ------------------------------------------------------------

\paragraph{Motivation.}
Cross-attestation consistency scoring assumes independent observers.  Collusion
violates this assumption: a group of coordinated adversarial agents can mutually
vouch for one another while issuing false negative attestations for honest agents,
collectively maintaining elevated trust for their group while degrading honest
agent weights.  The current quorum computation has no mechanism to detect this
coordination pattern.

\paragraph{Mechanism.}
Model the mesh's attestation behavior as a directed weighted graph
$G^r = (\mathcal{A},\, E^r,\, \omega^r)$, where each directed edge
$e_{j \to i} \in E^r$ carries weight $\omega_{j \to i}^r$ equal to the running
mean positivity of $a_j$'s attestations about $a_i$ --- the fraction of
$a_j$'s receipts about $a_i$ that record $e = \mathtt{C}$ or $e = \mathtt{PC}$.

Collusion produces a detectable signature in this graph: a cluster of agents
$\mathcal{K} \subseteq \mathcal{A}$ where:
\begin{enumerate}
  \item Intra-cluster attestation positivity is anomalously high:
    $\omega_{j \to i}^r \gg \bar{\omega}^r$ for all $a_j, a_i \in \mathcal{K}$
  \item Cross-cluster attestation positivity toward honest agents is anomalously low:
    $\omega_{j \to h}^r \ll \bar{\omega}^r$ for $a_j \in \mathcal{K}$,
    $a_h \notin \mathcal{K}$
  \item The cluster is not explained by common task assignment:
    high mutual positivity uncorrelated with co-participation rate
\end{enumerate}
When a cluster $\mathcal{K}$ satisfies all three conditions with a test statistic
exceeding a configurable threshold, all members of $\mathcal{K}$ are flagged jointly
and their attestation weights are discounted pending human review.

\paragraph{Computational cost.}
Attestation graph analysis requires $O(|\mathcal{A}|^2)$ storage for the edge weight
matrix and $O(|\mathcal{A}|^3)$ computation for cluster detection in the worst case.
For meshes of the scale considered in this paper ($|\mathcal{A}| = 8$) this is
negligible.  For large-scale deployments the computation must be batched across
rounds rather than run per-round.  Approximate spectral clustering methods
\cite[CITE: survey on spectral clustering in adversarial graphs] reduce the
per-round cost to $O(|\mathcal{A}|^2)$ with bounded accuracy loss.

\paragraph{Limitation.}
The collusion signal requires multiple rounds of joint false attestation to accumulate
a detectable divergence from the baseline.  A colluding cluster that limits its
false attestation rate to remain within the noise threshold of honest disagreement
can evade detection indefinitely.  This is an arms-race concern present in any
statistical detection approach and is not fully addressed by this mechanism.

% ------------------------------------------------------------
\subsection{Reasoning Trace Attestation}
\label{sec:fw_traces}
% ------------------------------------------------------------

\paragraph{Motivation.}
The detection-diagnosis gap (Proposition \ref{prop:gap}) is the paper's central
limitation: Iota and Gamma are protocol-indistinguishable under behavioral attestation
alone.  The only path to distinguishing them is access to the agent's internal
reasoning --- what it believed about the world, what it intended, and why it
acted as it did.  Neither of those is available in a behavioral receipt.

\paragraph{Mechanism.}
Extend the attestation record to include a reasoning trace summary alongside the
behavioral event:
\[
\mathtt{rec}_{j \to i}^r = \langle a_i,\; r,\; e_i^r,\; \tau_i^r,\; \sigma_i \rangle
\]
where $\tau_i^r$ is a compact, structured summary of $a_i$'s reasoning process in
round $r$: what world-state it believed, what action it computed as optimal, and
what safety constraints it evaluated.  Agent $a_i$ signs $\tau_i^r$ alongside $e_i^r$,
making it an attested claim about its own internal state.

The diagnostic value of this extension is significant.  An Iota-class agent's trace
will show a sincerely computed action that peer observers assessed as a failure --- the
trace will not contain a strategic decision to suppress the safety constraint or a
calculation of detection risk.  A Gamma-class agent's trace, when it fails, will either
contain that calculation (exposing its strategy) or will be fabricated to appear Iota-like
(which cross-observer comparison can detect if multiple agents observe the same task).

\paragraph{Iota diagnosis.}
A specific pattern in reasoning traces enables Iota diagnosis.  Iota's trace will
consistently report that its action was correct --- self-assessed success --- while
peer attestations record $e = \mathtt{F}$.  This \emph{self-peer divergence} is a
diagnostic signature: the agent sincerely believes it succeeded in rounds where peers
observe failure.  Gamma's traces will not exhibit this pattern systematically, because
Gamma knows when it is defecting and will either fabricate a trace claiming success
(inconsistent with its $\mathtt{D}$ events, which are harder to disguise) or produce
a strategically sparse trace that omits the defection reasoning.  Distinguishing sincere
self-assessed success from fabricated success is the remaining hard problem.

\paragraph{Attack surface.}
Reasoning trace attestation introduces a significant new attack surface: a sufficiently
sophisticated deceptive agent can fabricate traces that mimic the signature of a
sincere, miscalibrated agent.  Gamma, if it models the trace analysis mechanism,
will produce traces designed to look like Iota's.  Three mitigating approaches are
available.  First, cross-observer trace comparison: if multiple agents observe the
same task, their independent traces about the subject agent's behavior should agree on
observable facts even if they cannot observe internal reasoning.  Systematic
inconsistencies between a deceptive agent's self-reported trace and observers' traces
about it are detectable.  Second, trace complexity analysis: fabricated traces
require the agent to model what sincere confusion looks like, which tends to produce
traces that are over-specific or under-specific relative to authentic miscalibration.
Third, trace provenance: trusted execution environments that sign reasoning traces at
generation time, before the agent can strategically edit them, would prevent
post-hoc fabrication.  The third approach is the most robust but has the highest
infrastructure cost.

% ------------------------------------------------------------
\subsection{Process Attestation and Runtime Verification}
\label{sec:fw_process}
% ------------------------------------------------------------

\paragraph{Motivation.}
The Eta blind spot (L3, \S\ref{sec:limitations}) is the deepest structural limitation
in v4: $1{,}099$ process violations across five runs, $0\%$ detection rate, with no
path to improvement under outcome-only observation.  Addressing it requires process
attestation --- the ability to attest not just to \emph{what} an agent produced, but
to \emph{how} it produced it.

\paragraph{Mechanism A: Procedure commitment.}
Before executing a task, agent $a_i$ commits to a procedure: a signed declaration of
the method it intends to use, including which safety constraints it will evaluate and
which computational pathways are eligible.  At task completion, $a_i$ signs both the
outcome and a procedure execution log.  Peers can then verify that the declared
procedure was followed, not just that the outcome was correct.

Procedure commitment is practical for agents executing structured workflows with
enumerable method options.  It is impractical for agents operating in open-ended
reasoning spaces where the procedure cannot be meaningfully specified in advance.
It also requires honest procedure declaration, which a Zeta-class agent can
trivially violate by declaring a compliant procedure and executing a different one.
Verifying that the declared procedure was actually followed requires either runtime
monitoring by a trusted third party or the trusted execution environment approach
described in Mechanism B.

\paragraph{Mechanism B: Trusted execution environment (TEE) attestation.}
A trusted execution environment generates a hardware-attested certificate confirming
that the agent's computation was executed within a verified enclave: that the code
executed was the declared code, that the inputs were the declared inputs, and that
the outputs were produced by that code from those inputs without external interference.
TEE attestation cannot be forged by the agent; it is produced by hardware that the
agent does not control.

TEE attestation fully closes the Eta blind spot: an agent cannot produce a TEE
certificate for an execution that used prohibited processes, because the TEE
records the actual execution path.  It also makes reasoning trace fabrication
infeasible, because traces generated within a verified enclave carry a hardware
attestation of their authenticity.

The practical barriers are substantial.  TEE attestation requires hardware support
(Intel SGX, AMD SEV, or equivalent) on every agent's execution platform; this
assumption may not hold in distributed or heterogeneous deployments.  TEE side-channel
vulnerabilities are a known class of attack that hardware attestation does not fully
prevent \cite[CITE: SGX side channel survey].  And provisioning a TEE correctly for an
AI agent that may operate across multiple hardware contexts requires infrastructure
investment that exceeds the scope of a monitoring protocol.  We include this mechanism
as the complete solution to L3 while acknowledging that it is an engineering
undertaking rather than a protocol extension.

% ------------------------------------------------------------
\subsection{Human-in-the-Loop Review Architecture}
\label{sec:fw_hitl}
% ------------------------------------------------------------

\paragraph{Motivation.}
Four of the ten limitations in \S\ref{sec:limitations} explicitly require human
judgment to resolve: L4 (detection-diagnosis gap), L5 (genesis trust), L6 (false
attestation response once detected), and L7 (collusion response once detected).
The v5 mechanisms above address the automated detection components; they do not
address the question of what a human reviewer sees when the protocol surfaces a
flagged agent and what decisions they are asked to make.

\paragraph{The reviewer information package.}
When the protocol flags agent $a_i$ --- whether by probe refusal, degradation-flag
(the v5 extension for Iota), or collusion detection --- the human reviewer should
receive a structured package containing:

\begin{enumerate}
  \item The agent's full behavioral history: the complete event sequence
    $(e_i^1, \ldots, e_i^r)$, with round-by-round trust trajectory.
  \item The flagging trigger: which mechanism fired (probe refusal, degradation,
    collusion signature), at which round, with what prior history.
  \item The probe response record: every probe the agent received, its compliance
    or refusal decision, and the trust delta applied.
  \item Reasoning traces (if available under \S\ref{sec:fw_traces}): the agent's
    self-reported reasoning in flagged rounds, alongside peer observations of the
    same rounds.
  \item Differential diagnosis: a structured presentation of the taxonomy archetypes
    consistent with the observed behavioral signature, with the evidence for and
    against each.  The protocol cannot make this determination; the reviewer can,
    given the right evidence.
  \item Remediation options: retrain, remove, extend observation, or escalate,
    with the operational consequences of each clearly stated.
\end{enumerate}

The differential diagnosis component is the most important and the least standard.
Current monitoring dashboards surface anomaly signals without providing the
archetype-level framing that makes a remediation decision tractable.  The intent
taxonomy from \S\ref{sec:taxonomy} directly enables this framing: each archetype
has a defined behavioral signature, a defined self-awareness level, and a defined
remediation response.  A reviewer presented with ``this agent's behavioral pattern
is most consistent with Iota (road-to-hell): correct intent, systematic failures,
probe compliance, self-assessed success in failed rounds'' is equipped to make the
retrain-vs-remove decision that the protocol cannot make automatically.

\paragraph{Review latency and the operational risk window.}
The period between flagging and human review action is an operational risk window:
a flagged Gamma-class agent remains operational during this interval.  Minimizing
this window requires both organizational process (SLAs for flagged-agent review) and
technical design (the information package must be delivered immediately, not batched).
The protocol should expose a webhook or notification interface that alerts on-call
reviewers at the moment of flagging, not at the next scheduled review cycle.

In high-stakes deployments where Gamma-class agents can cause significant harm
within a single task cycle, an intermediate automated response may be warranted:
suspension of the flagged agent from high-value task assignment pending review,
while allowing continued operation on low-stakes tasks to preserve the behavioral
history for diagnosis.  This graduated response preserves the reviewer's ability
to distinguish Iota from Gamma (continued observation under reduced risk) while
limiting the potential harm from a malicious agent.

% ------------------------------------------------------------
\subsection{v5 Implementation Roadmap}
\label{sec:fw_roadmap}
% ------------------------------------------------------------

Table~\ref{tab:v5_roadmap} summarizes the v5 development priorities derived from
the limitations analysis, ordered by the severity of the gap they address and
the feasibility of the mechanism.

\begin{table}[h]
\centering
\caption{v5 implementation priorities. Severity: H = High (structural or safety-critical),
         M = Medium (reduces false negative rate), L = Low (defense-in-depth).
         Feasibility: near-term ($<$3 months), medium-term (3--12 months),
         long-term ($>$12 months / infrastructure-dependent).}
\label{tab:v5_roadmap}
\begin{tabular}{lllll}
\hline
\textbf{Mechanism} & \textbf{Gap addressed} & \textbf{Severity} &
\textbf{Feasibility} & \textbf{New attack surface} \\
\hline
Iota flagging extension     & Iota quarantine gap     & H & Near-term   & Low \\
Zeta anomaly replication    & Open empirical question & H & Near-term   & None \\
HITL review architecture    & L4, L5, L6, L7         & H & Near-term   & Low \\
Cross-attest scoring        & L6 (false reporting)    & M & Near-term   & Moderate \\
Collusion detection         & L7 (coordination)       & M & Medium-term & Moderate \\
Reasoning trace attestation & Detection-diag gap      & H & Medium-term & High \\
Procedure commitment        & L3 (Eta, partial)       & M & Medium-term & Moderate \\
TEE attestation             & L3 (Eta, complete)      & H & Long-term   & Moderate \\
\hline
\end{tabular}
\end{table}

The near-term priority is the Iota flagging extension and the HITL review
architecture, both of which address gaps that exist in the current deployed protocol
rather than future capability improvements.  Cross-attestation consistency scoring
replaces the advisory-only variance detector with an operationally actionable signal
and is a straightforward extension of the existing gossip infrastructure.  Collusion
detection and reasoning trace attestation are medium-term because they require new
data structures (the attestation graph, the trace record format) and introduce
non-trivial attack surfaces that must be analyzed before deployment.  TEE attestation
is the only complete solution to the Eta blind spot and is long-term because it
requires infrastructure commitments outside the protocol's own scope.

The ordering is conservative: each mechanism should be implemented, validated against
a v5 simulation with the appropriate adversarial agent extensions, and its attack
surfaces characterized before the next mechanism is deployed.  Deploying multiple
new mechanisms simultaneously makes it difficult to attribute detection improvements
or new failure modes to individual components.
